---
title: "Modeling Left Ventricular Diastolic Dysfunction and Cardiopulmonary Fitness in HCM"
author: "Mark E. Pepin, MD, PhD, MS, FESC"
date: "2026-03-25"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-tools: true
    code-fold: true
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 6.9,
  fig.height = 4.5,
  dpi = 300,
  dev = "cairo_pdf",
  dev.args = list(family = "Arial")
)

# ── Package loading ──────────────────────────────────────────────────────────
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(patchwork)
  library(splines)
  library(lme4)
  library(mgcv)
  library(broom)
  library(survival)
  library(scales)
  library(stringr)
  library(openxlsx)
  library(gtsummary)
  library(flextable)
  library(ggcorrplot)
  library(viridis)
  library(ggrepel)
  library(gamlss)
})

# ── JACC color palette (colorblind-safe) ─────────────────────────────────────
jacc_cols <- c(
  navy    = "#2C3E50",
  red     = "#E74C3C",
  blue    = "#3498DB",
  green   = "#27AE60",
  orange  = "#F39C12",
  gray    = "#7F8C8D",
  purple  = "#8E44AD"
)

# ── JACC publication theme ───────────────────────────────────────────────────
theme_jacc <- function(base_size = 9) {
  theme_classic(base_size = base_size, base_family = "Arial") +
    theme(
      plot.title       = element_text(face = "bold", size = base_size + 1, hjust = 0.5),
      plot.subtitle    = element_text(size = base_size, hjust = 0.5, color = "grey40"),
      axis.title       = element_text(face = "bold"),
      axis.text        = element_text(color = "black"),
      axis.line        = element_line(color = "black", linewidth = 0.4),
      axis.ticks       = element_line(color = "black", linewidth = 0.3),
      panel.grid       = element_blank(),
      strip.background = element_rect(fill = "grey94", color = NA),
      strip.text       = element_text(face = "bold", size = base_size),
      legend.position  = "top",
      legend.title     = element_blank(),
      legend.text      = element_text(size = base_size - 1),
      plot.margin      = margin(4, 8, 4, 4)
    )
}

# ── Export helper ────────────────────────────────────────────────────────────
save_jacc <- function(plot, path, w = 6.9, h = 4.5) {
  ggsave(filename = path, plot = plot, device = cairo_pdf,
         width = w, height = h, units = "in", dpi = 600)
}

# ── Age-residualization helper (single definition) ───────────────────────────
residualize_on_age <- function(df, idx) {
  df <- df %>% mutate(.row_id = row_number())
  dsub <- df %>% filter(!is.na(.data[[idx]]), !is.na(age))
  if (nrow(dsub) < 10) {
    df[[paste0(idx, "_resid")]] <- NA_real_
    return(df %>% select(-.row_id))
  }
  fit <- lm(as.formula(paste0(idx, " ~ ns(age, df = 3)")), data = dsub)
  dsub$.resid <- resid(fit)
  df <- df %>% left_join(dsub %>% select(.row_id, .resid), by = ".row_id")
  df[[paste0(idx, "_resid")]] <- df$.resid
  df %>% select(-.row_id, -.resid)
}

# ── Output directory ─────────────────────────────────────────────────────────
dir.create("../2_Output", recursive = TRUE, showWarnings = FALSE)
```

# Data Import and Cohort Assembly

## Import and clean CPX data

The initial phase imports raw Cardiopulmonary Exercise Testing (CPET) data, assigns a stable patient identifier (ID) per MRN, retains patients with serial testing ($\geq 2$ sessions), restricts to HCM with adequate effort (peak RER $\geq 1.0$), and derives biometric indices (BSA, BMI, LBMI).

```{r cpx_cleaning}
CPX <- read.xlsx("../1_Input/CPX_input.xlsx", sheet = "CPX", detectDates = TRUE)

CPX_with_id <- CPX %>%
  group_by(MRN) %>%
  mutate(ID = cur_group_id()) %>%
  ungroup()

CPX_serial <- CPX_with_id %>%
  group_by(ID) %>%
  filter(n() > 1) %>%
  ungroup()

CPX_hcm_effort <- CPX_serial %>%
  filter(ID != "1380") %>%
  filter(MCl3 == "HCM", pk.RER >= 1.0)

# Cohort flow tracking
filter_flow <- tibble(
  stage      = c("Initial CPX import", "Assigned stable patient ID",
                 "Retained serial CPX only", "HCM diagnosis + RER >= 1.0"),
  n_tests    = c(nrow(CPX), nrow(CPX_with_id), nrow(CPX_serial), nrow(CPX_hcm_effort)),
  n_patients = c(n_distinct(CPX$MRN), n_distinct(CPX_with_id$ID),
                 n_distinct(CPX_serial$ID), n_distinct(CPX_hcm_effort$ID))
)

# Biometric calculations
CPX_clean <- CPX_hcm_effort %>%
  mutate(
    BSA  = sqrt((`height.(cm)` * `Weight.(kg)`) / 3600),
    LBMI = LBM.NHANES.no.race / (`height.(cm)` / 100)^2,
    BMI  = `Weight.(kg)` / ((`height.(cm)` / 100)^2)
  ) %>%
  group_by(ID) %>%
  arrange(cpx_test_date, .by_group = TRUE) %>%
  mutate(days_from_baseline = as.numeric(cpx_test_date - first(cpx_test_date))) %>%
  ungroup()

CPX_clean <- CPX_clean %>%
  select(
    ID, MRN, cpx_test_date, age, Sex = `Sex:.M(0)/F(1)`, BMI, BSA,
    LBM = LBM.NHANES.no.race, LBMI, HCM_Phenotype = MCl4,
    CPX.sequential, days_from_baseline, pk.RER,
    VO2_FRIEND_PP   = `FRIEND1.%.predicted.VO2`,
    VO2_WASSERMAN_PP = `Wasserman.%predicted.VO2.(2005)`,
    VO2_FRIEND2_PP  = `FRIEND2.%predicted.VO2`,
    HRmax_PP = `%predicted.HR.FRIEND`,
    HRR = `HR.recovery.(1min)`,
    VeVco2_slope = `ve/vco2.slope`
  ) %>%
  filter(
    (is.na(VO2_FRIEND2_PP) | (VO2_FRIEND2_PP >= 0 & VO2_FRIEND2_PP <= 200)) &
    (is.na(VO2_FRIEND_PP)  | (VO2_FRIEND_PP >= 0 & VO2_FRIEND_PP <= 200))
  )

cat(sprintf("CPX tests (HCM, adequate effort): %d tests, %d patients\n",
            nrow(CPX_clean), n_distinct(CPX_clean$ID)))
```

## Comorbidity exclusions

Patients with documented CAD, COPD, or interstitial lung disease are excluded so that exercise limitations are attributable to HCM-specific pathology.

```{r comorbidities}
Comorbidities_table <- read.xlsx("../1_Input/CPX_input.xlsx",
                                  sheet = "co-morbidities", detectDates = TRUE)

Comorbidities_to_filter <- Comorbidities_table %>%
  select(MRN, cad_pre_test, copd_pre_test, interstitial_lung_dz_pre_test) %>%
  filter(
    (cad_pre_test != 1 | is.na(cad_pre_test)) &
    (copd_pre_test != 1 | is.na(copd_pre_test)) &
    (interstitial_lung_dz_pre_test != 1 | is.na(interstitial_lung_dz_pre_test))
  )

CPX_comorbidity_filtered <- CPX_clean %>%
  filter(MRN %in% Comorbidities_to_filter$MRN)

filter_flow <- bind_rows(filter_flow, tibble(
  stage      = "Excluded CAD/COPD/ILD",
  n_tests    = nrow(CPX_comorbidity_filtered),
  n_patients = n_distinct(CPX_comorbidity_filtered$ID)
))
```

## Merge echocardiographic data

Each CPET is aligned with the closest prior resting echocardiogram within a configurable window (default 365 days).

```{r echo_merge}
echo_raw <- read.xlsx("../1_Input/mark_extract_20260316.xlsx", sheet = 1, detectDates = TRUE)

echo_vars_core <- c(
  "e_e_ave", "e_e_lat", "e_e_med", "mv_a_dur", "mv_a_point", "mv_dec_time",
  "mv_e_a", "tr_max_vel", "la_vol_index", "med_peak_e_vel", "lat_peak_e_vel",
  "ef_modsp4", "la_vol", "pulm_dias_vel", "pulm_sys_vel"
)

echo_all <- echo_raw %>%
  mutate(MRN = as.character(mrn), echo_date = as.Date(echo_date)) %>%
  select(MRN, echo_date, all_of(echo_vars_core)) %>%
  mutate(across(all_of(echo_vars_core), ~ as.numeric(as.character(.))))

# Closest prior echo within window
align_window_days <- 365

cpx_aligned <- CPX_comorbidity_filtered %>%
  mutate(MRN = as.character(MRN), cpx_test_date = as.Date(cpx_test_date))

cpx_echo <- cpx_aligned %>%
  left_join(echo_all, by = "MRN", relationship = "many-to-many") %>%
  mutate(delta_days = as.numeric(cpx_test_date - echo_date)) %>%
  filter(!is.na(delta_days), delta_days >= 0, delta_days <= align_window_days) %>%
  group_by(ID, CPX.sequential, cpx_test_date) %>%
  slice_min(order_by = delta_days, n = 1, with_ties = FALSE) %>%
  ungroup()

# Prepare derived variables
cpx_echo <- cpx_echo %>%
  mutate(
    E_vel = coalesce(med_peak_e_vel, lat_peak_e_vel),
    Sex   = factor(Sex, levels = c(0, 1), labels = c("Male", "Female"))
  )

# Update filter flow
filter_flow <- bind_rows(filter_flow, tibble(
  stage      = "Echo aligned (within 365d)",
  n_tests    = nrow(cpx_echo),
  n_patients = n_distinct(cpx_echo$ID)
))

# Baseline and longitudinal cohorts
num_vars <- c("VO2_FRIEND2_PP", "VO2_FRIEND_PP", "VO2_WASSERMAN_PP", "VeVco2_slope",
              "pk.RER", "age", "BMI", "LBMI", "e_e_ave", "la_vol_index", "tr_max_vel",
              "med_peak_e_vel", "lat_peak_e_vel", "ef_modsp4")

baseline_df <- cpx_echo %>%
  group_by(ID) %>% arrange(days_from_baseline) %>% slice(1) %>% ungroup() %>%
  mutate(across(any_of(num_vars), ~ as.numeric(as.character(.))))

long_df <- cpx_echo %>%
  filter(!is.na(VO2_FRIEND2_PP), !is.na(days_from_baseline)) %>%
  mutate(across(any_of(num_vars), ~ as.numeric(as.character(.))),
         time_yrs = days_from_baseline / 365.25)

cat(sprintf("Baseline cohort: %d patients\nLongitudinal observations: %d\n",
            nrow(baseline_df), nrow(long_df)))
```

## Import clinical outcomes

```{r outcomes_import}
outcomes_raw <- read.xlsx("../1_Input/CPX_input.xlsx", sheet = "Outcomes_2025", detectDates = TRUE)

outcomes <- outcomes_raw %>%
  mutate(MRN = as.character(MRN)) %>%
  select(MRN,
         death, death_date_clean,
         post_acute_heart_failure, post_acute_heart_failure_date, post_acute_heart_failure_yrs,
         post_chronic_heart_failure, post_chronic_heart_failure_date, post_chronic_heart_failure_yrs,
         post_heart_transplant, post_heart_transplant_date, post_heart_transplant_yrs,
         post_afib_flut, post_afib_flut_date, post_afib_flut_yrs,
         last_enc_date) %>%
  mutate(across(c(death, post_acute_heart_failure, post_chronic_heart_failure,
                  post_heart_transplant, post_afib_flut), ~ as.numeric(as.character(.)))) %>%
  mutate(across(c(post_acute_heart_failure_yrs, post_chronic_heart_failure_yrs,
                  post_heart_transplant_yrs, post_afib_flut_yrs), ~ as.numeric(as.character(.))))

# Composite HF endpoint: acute HF, chronic HF, transplant, or death
outcomes <- outcomes %>%
  mutate(
    hf_composite = pmax(
      coalesce(post_acute_heart_failure, 0),
      coalesce(post_chronic_heart_failure, 0),
      coalesce(post_heart_transplant, 0),
      coalesce(death, 0),
      na.rm = TRUE
    ),
    hf_composite_yrs = pmin(
      coalesce(post_acute_heart_failure_yrs, Inf),
      coalesce(post_chronic_heart_failure_yrs, Inf),
      coalesce(post_heart_transplant_yrs, Inf),
      na.rm = TRUE
    ),
    hf_composite_yrs = ifelse(is.infinite(hf_composite_yrs), NA_real_, hf_composite_yrs)
  )

# Merge to baseline
baseline_df <- baseline_df %>%
  left_join(outcomes %>% select(MRN, hf_composite, hf_composite_yrs,
                                 post_afib_flut, post_afib_flut_yrs,
                                 death, last_enc_date),
            by = "MRN")

cat(sprintf("Patients with outcome data: %d\nComposite HF events: %d\n",
            sum(!is.na(baseline_df$hf_composite)),
            sum(baseline_df$hf_composite == 1, na.rm = TRUE)))
```

## Cohort flow diagram

```{r flow_diagram, fig.width=6.9, fig.height=3.8, results='asis'}
filter_flow <- filter_flow %>%
  mutate(
    stage = factor(stage, levels = stage),
    removed = lag(n_tests, default = first(n_tests)) - n_tests
  )

p_flow <- filter_flow %>%
  mutate(stage = factor(stage, levels = rev(levels(stage)))) %>%
  ggplot(aes(y = stage, x = n_tests)) +
  geom_col(fill = jacc_cols["navy"], width = 0.65, color = "black", linewidth = 0.2) +
  geom_text(aes(label = paste0(comma(n_tests), " tests | ", comma(n_patients), " pts")),
            hjust = 0, nudge_x = 15, size = 2.8, color = "black") +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.25))) +
  labs(title = "Cohort Assembly", x = "CPX Tests", y = NULL) +
  theme_jacc()

p_flow
save_jacc(p_flow, "../2_Output/Cohort_Filtering_Workflow.pdf", w = 6.9, h = 3.8)
```

# ASE 2025 Diastolic Dysfunction Grading

The 2025 ASE guidelines propose a hierarchical framework integrating E/e' (>14), Ar-A duration difference ($\geq$ 30 ms), left atrial volume index (>34 mL/m$^2$), and peak tricuspid regurgitation velocity (>2.8 m/s). Pulmonary venous Ar duration was not available in this dataset; grading proceeds with 3 of 4 criteria.

```{r ase_grading, results='asis'}
grade_ase2025 <- function(e_e_ave, la_vol_index, tr_max_vel) {
  # Count abnormal criteria (excluding Ar-A which is unavailable)
  # Note: tr_max_vel is stored in cm/s; ASE cutoff is >2.8 m/s = >280 cm/s
  abn_ee   <- !is.na(e_e_ave)     & e_e_ave > 14
  abn_lavi <- !is.na(la_vol_index) & la_vol_index > 34
  abn_trv  <- !is.na(tr_max_vel)   & tr_max_vel > 280

  n_available <- (!is.na(e_e_ave)) + (!is.na(la_vol_index)) + (!is.na(tr_max_vel))
  n_abnormal  <- abn_ee + abn_lavi + abn_trv

  grade <- case_when(
    n_available < 2                         ~ "Indeterminate",
    abn_ee & abn_lavi & abn_trv             ~ "Grade III",
    abn_ee & (abn_lavi | abn_trv)           ~ "Grade II",
    abn_ee & n_abnormal == 1                ~ "Grade I",
    !abn_ee & n_abnormal >= 2               ~ "Grade I",
    !abn_ee & n_abnormal <= 1               ~ "Normal",
    TRUE                                     ~ "Indeterminate"
  )

  factor(grade, levels = c("Normal", "Grade I", "Grade II", "Grade III", "Indeterminate"),
         ordered = FALSE)
}

baseline_df <- baseline_df %>%
  mutate(dd_grade = grade_ase2025(e_e_ave, la_vol_index, tr_max_vel))

long_df <- long_df %>%
  left_join(
    baseline_df %>% select(ID, dd_grade),
    by = "ID"
  )

cat("**ASE 2025 DD Grade Distribution (Baseline):**\n\n")
table(baseline_df$dd_grade, useNA = "ifany")
```

```{r ase_figure, fig.width=6.9, fig.height=4.0, results='asis'}
# Panel A: DD grade distribution
df_grade <- baseline_df %>% filter(dd_grade != "Indeterminate")

p_bar <- df_grade %>%
  count(dd_grade) %>%
  ggplot(aes(x = dd_grade, y = n, fill = dd_grade)) +
  geom_col(width = 0.7, color = "black", linewidth = 0.2) +
  geom_text(aes(label = n), vjust = -0.4, size = 3) +
  scale_fill_manual(values = c("Normal" = jacc_cols["blue"],
                                "Grade I" = jacc_cols["green"],
                                "Grade II" = jacc_cols["orange"],
                                "Grade III" = jacc_cols["red"])) +
  labs(title = "DD Grade Distribution", x = "ASE 2025 Grade", y = "Patients") +
  theme_jacc() + theme(legend.position = "none")

# Panel B: Peak VO2 by DD grade
p_box <- df_grade %>%
  filter(!is.na(VO2_FRIEND2_PP)) %>%
  ggplot(aes(x = dd_grade, y = VO2_FRIEND2_PP, fill = dd_grade)) +
  geom_boxplot(width = 0.6, outlier.size = 0.8, linewidth = 0.3) +
  scale_fill_manual(values = c("Normal" = jacc_cols["blue"],
                                "Grade I" = jacc_cols["green"],
                                "Grade II" = jacc_cols["orange"],
                                "Grade III" = jacc_cols["red"])) +
  labs(title = bquote(bold("Peak" ~ VO[2] ~ "by DD Grade")),
       x = "ASE 2025 Grade",
       y = bquote("Peak" ~ VO[2] ~ "(% Predicted, FRIEND 2.0)")) +
  theme_jacc() + theme(legend.position = "none")

# Panel C: VE/VCO2 by DD grade
p_vevco2_grade <- df_grade %>%
  filter(!is.na(VeVco2_slope)) %>%
  ggplot(aes(x = dd_grade, y = VeVco2_slope, fill = dd_grade)) +
  geom_boxplot(width = 0.6, outlier.size = 0.8, linewidth = 0.3) +
  scale_fill_manual(values = c("Normal" = jacc_cols["blue"],
                                "Grade I" = jacc_cols["green"],
                                "Grade II" = jacc_cols["orange"],
                                "Grade III" = jacc_cols["red"])) +
  labs(title = bquote(bold(VE/VCO[2] ~ "Slope by DD Grade")),
       x = "ASE 2025 Grade",
       y = bquote(VE/VCO[2] ~ "Slope")) +
  theme_jacc() + theme(legend.position = "none")

fig_ase <- (p_bar | p_box | p_vevco2_grade) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "bold", size = 11))

fig_ase
save_jacc(fig_ase, "../2_Output/Figure_ASE2025_DDGrade.pdf", w = 6.9, h = 4.0)
```

```{r ase_stratified_table, results='asis'}
tbl_ase <- baseline_df %>%
  filter(dd_grade != "Indeterminate") %>%
  select(dd_grade, age, Sex, BMI, LBMI, HCM_Phenotype,
         VO2_FRIEND2_PP, VO2_WASSERMAN_PP, VeVco2_slope, HRR,
         e_e_ave, la_vol_index, tr_max_vel, ef_modsp4) %>%
  tbl_summary(
    by = dd_grade,
    label = list(
      age            ~ "Age, years",
      Sex            ~ "Sex",
      BMI            ~ "BMI, kg/m\u00B2",
      LBMI           ~ "LBMI, kg/m\u00B2",
      HCM_Phenotype  ~ "HCM Phenotype",
      VO2_FRIEND2_PP ~ "Peak VO\u2082 (FRIEND 2.0 %pred)",
      VO2_WASSERMAN_PP ~ "Peak VO\u2082 (Wasserman %pred)",
      VeVco2_slope   ~ "VE/VCO\u2082 Slope",
      HRR            ~ "HR Recovery (1 min)",
      e_e_ave        ~ "E/e' (average)",
      la_vol_index   ~ "LAVI, mL/m\u00B2",
      tr_max_vel     ~ "TR Vmax, m/s",
      ef_modsp4      ~ "LVEF, %"
    ),
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 1,
    missing = "no"
  ) %>%
  add_p() %>%
  bold_labels() %>%
  modify_caption("**Table: Patient Characteristics Stratified by ASE 2025 DD Grade**")

tbl_ase

tbl_ase %>% as_flex_table() %>%
  save_as_docx(path = "../2_Output/Table_ASE2025_Stratified.docx")
```

# GAMLSS / SEAMLSS Continuous Modeling

We employ a Semiparametric Additive Model for Location, Scale, and Shape (GAMLSS) to model the continuous relationship between diastolic indices and Peak VO$_2$. The SEAMLSS framework derives Z-scores via $u(x) = \ln(y / \mu(x))$, where $\mu(x)$ is the fitted conditional mean, providing a stable relative metric that avoids the global distortion of Box-Cox (LMS) transformations.

```{r gamlss_modeling}
# Fit GAMLSS for each key diastolic index against Peak VO2
diastolic_key <- c("e_e_ave", "la_vol_index", "tr_max_vel")
diastolic_labels <- c("e_e_ave" = "E/e' (average)",
                       "la_vol_index" = "LAVI (mL/m\u00B2)",
                       "tr_max_vel" = "TR Vmax (m/s)")

gamlss_results <- list()

for (idx in diastolic_key) {
  df_fit <- baseline_df %>%
    filter(!is.na(.data[[idx]]), !is.na(VO2_FRIEND2_PP), !is.na(age), !is.na(LBMI)) %>%
    mutate(Sex_num = as.numeric(Sex == "Female"))

  if (nrow(df_fit) < 30) next

  # GAMLSS requires no NAs in the data frame — subset to used columns only
  gamlss_cols <- c("VO2_FRIEND2_PP", idx, "age", "Sex_num", "LBMI")
  df_gamlss <- df_fit[, gamlss_cols, drop = FALSE] %>% na.omit()
  if (nrow(df_gamlss) < 30) next

  tryCatch({
    frm_mu <- as.formula(paste0("VO2_FRIEND2_PP ~ cs(", idx, ", df=3) + cs(age, df=3) + Sex_num + LBMI"))
    m <- gamlss(
      frm_mu,
      sigma.formula = ~ cs(age, df=3),
      data = df_gamlss,
      family = NO(),
      control = gamlss.control(n.cyc = 100, trace = FALSE)
    )

    # SEAMLSS Z-score: u(x) = ln(y / mu(x))
    mu_hat <- fitted(m, what = "mu")
    sigma_hat <- fitted(m, what = "sigma")
    df_gamlss$seamlss_z <- log(df_gamlss$VO2_FRIEND2_PP / mu_hat)
    df_gamlss$gamlss_z  <- (df_gamlss$VO2_FRIEND2_PP - mu_hat) / sigma_hat
    df_gamlss$mu_hat    <- mu_hat

    # Merge back to df_fit (which has ID and other columns)
    df_out <- df_fit
    df_out$seamlss_z <- NA_real_; df_out$gamlss_z <- NA_real_; df_out$mu_hat <- NA_real_
    match_idx <- match(rownames(df_gamlss), rownames(df_fit))
    df_out$seamlss_z[match_idx] <- df_gamlss$seamlss_z
    df_out$gamlss_z[match_idx]  <- df_gamlss$gamlss_z
    df_out$mu_hat[match_idx]    <- df_gamlss$mu_hat

    gamlss_results[[idx]] <- list(model = m, data = df_out)
  }, error = function(e) {
    message(sprintf("GAMLSS for %s failed: %s", idx, e$message))
  })
}
```

```{r gamlss_figures, fig.width=6.9, fig.height=6.5}
if (length(gamlss_results) > 0) {
  # Panel plots: observed vs fitted centile curves
  centile_plots <- lapply(names(gamlss_results), function(idx) {
    res <- gamlss_results[[idx]]
    df  <- res$data

    ggplot(df, aes(x = .data[[idx]], y = VO2_FRIEND2_PP)) +
      geom_point(alpha = 0.3, size = 1, color = jacc_cols["gray"]) +
      geom_smooth(aes(y = mu_hat), color = jacc_cols["navy"], linewidth = 1,
                  method = "loess", se = FALSE) +
      geom_smooth(method = "loess", color = jacc_cols["red"], linewidth = 0.7,
                  linetype = "dashed", se = TRUE, alpha = 0.1) +
      labs(title = diastolic_labels[idx],
           x = diastolic_labels[idx],
           y = bquote("Peak" ~ VO[2] ~ "(% Pred)")) +
      theme_jacc()
  })

  # SEAMLSS Z-score distribution
  z_data <- bind_rows(lapply(names(gamlss_results), function(idx) {
    gamlss_results[[idx]]$data %>%
      select(seamlss_z) %>%
      mutate(index = diastolic_labels[idx])
  }))

  p_zdist <- ggplot(z_data, aes(x = seamlss_z, fill = index)) +
    geom_histogram(aes(y = after_stat(density)), bins = 30, alpha = 0.6, color = "white") +
    geom_density(aes(color = index), linewidth = 0.8) +
    facet_wrap(~ index, scales = "free") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(title = "SEAMLSS Z-Score Distribution",
         x = "Z-score: ln(observed / predicted)",
         y = "Density") +
    theme_jacc() + theme(legend.position = "none")

  # Combine
  n_plots <- length(centile_plots)
  top_row <- wrap_plots(centile_plots, nrow = 1)
  fig_gamlss <- top_row / p_zdist +
    plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(face = "bold", size = 11))

  fig_gamlss
  save_jacc(fig_gamlss, "../2_Output/Figure_GAMLSS_SEAMLSS.pdf", w = 6.9, h = 6.5)
}
```

```{r gamlss_table, results='asis'}
if (length(gamlss_results) > 0) {
  gamlss_summary <- bind_rows(lapply(names(gamlss_results), function(idx) {
    m <- gamlss_results[[idx]]$model
    # Capture summary output as a matrix
    s <- tryCatch({
      capture.output(sm <- summary(m))
      as.data.frame(sm) %>%
        tibble::rownames_to_column("term") %>%
        mutate(index = diastolic_labels[idx], AIC = AIC(m))
    }, error = function(e) {
      # Fallback: extract coefficients directly
      co <- coef(m)
      tibble(term = names(co), Estimate = co, index = diastolic_labels[idx], AIC = AIC(m))
    })
    s
  }))

  gamlss_summary %>%
    mutate(across(where(is.numeric), ~ round(., 4))) %>%
    flextable() %>%
    bold(part = "header") %>%
    theme_booktabs() %>%
    autofit() %>%
    set_caption("Table 2. GAMLSS Model Coefficients for Diastolic Indices vs Peak VO\u2082")

  write.csv(gamlss_summary,
            "../2_Output/Table2_GAMLSS_Coefficients.csv", row.names = FALSE)
}
```

# Cross-Sectional Nonlinear Analysis

Individual diastolic indices are modeled against Peak VO$_2$ using natural splines (df=3) adjusted for age, sex, and lean body mass index. Likelihood ratio tests compare each diastolic-augmented model to the clinical base model.

```{r cross_sectional_helpers, include=FALSE}
diastolic_indices <- c(
 "e_e_ave", "e_e_lat", "e_e_med", "mv_a_dur", "mv_a_point", "mv_dec_time",
 "mv_e_a", "tr_max_vel", "la_vol_index", "med_peak_e_vel", "lat_peak_e_vel"
)

label_map <- c(
  e_e_ave       = "E/e' (avg)",    e_e_lat       = "E/e' (lat)",
  e_e_med       = "E/e' (med)",    mv_a_dur      = "MV A dur",
  mv_a_point    = "MV A point",    mv_dec_time   = "MV decel time",
  mv_e_a        = "MV E/A",        tr_max_vel    = "TR Vmax",
  la_vol_index  = "LAVI",          med_peak_e_vel= "Med E vel",
  lat_peak_e_vel= "Lat E vel"
)

pred_grid <- function(var, data, fit, x_from = NULL, x_to = NULL) {
  rng <- range(data[[var]], na.rm = TRUE)
  if (!is.null(x_from)) rng[1] <- x_from
  if (!is.null(x_to))   rng[2] <- x_to
  grid <- tibble(
    x = seq(rng[1], rng[2], length.out = 150),
    age = median(data$age, na.rm = TRUE),
    Sex = factor(names(which.max(table(data$Sex))), levels = levels(data$Sex)),
    LBMI = median(data$LBMI, na.rm = TRUE)
  )
  names(grid)[1] <- var
  p <- predict(fit, newdata = grid, interval = "confidence")
  bind_cols(grid, as.data.frame(p)) %>% rename(x = !!sym(var))
}
```

## Primary outcome: Peak VO$_2$ (FRIEND 2.0 % predicted)

```{r cross_sectional_vo2, fig.width=6.9, fig.height=6.5, results='asis'}
vo2_models <- list()
lrt_results <- list()

for (idx in diastolic_indices) {
  df_idx <- baseline_df %>%
    filter(!is.na(.data[["VO2_FRIEND2_PP"]]), !is.na(age), !is.na(Sex),
           !is.na(LBMI), !is.na(.data[[idx]]))
  if (nrow(df_idx) < 20) next

  base <- lm(VO2_FRIEND2_PP ~ age + Sex + LBMI, data = df_idx)
  mod  <- lm(as.formula(paste0("VO2_FRIEND2_PP ~ age + Sex + LBMI + ns(", idx, ", df=3)")),
             data = df_idx)
  lrt  <- anova(base, mod)

  vo2_models[[idx]] <- list(df = df_idx, base = base, mod = mod, lrt = lrt)
  lrt_results[[idx]] <- tibble(
    index = idx, label = label_map[idx], n = nrow(df_idx),
    delta_AIC = AIC(mod) - AIC(base),
    F_stat = lrt$F[2], p_value = lrt$`Pr(>F)`[2],
    R2_base = summary(base)$r.squared,
    R2_full = summary(mod)$r.squared
  )
}

lrt_table <- bind_rows(lrt_results) %>% arrange(p_value)
write.csv(lrt_table, "../2_Output/Stats_CrossSectional_LRT_FRIEND2.csv", row.names = FALSE)
print(knitr::kable(lrt_table %>% mutate(across(where(is.numeric), ~ round(., 3))),
                   caption = "Likelihood Ratio Tests: Diastolic Indices vs Peak VO2 (FRIEND 2.0)"))

# Spline prediction curves
if (length(vo2_models) > 0) {
  pp <- bind_rows(lapply(names(vo2_models), function(idx) {
    m <- vo2_models[[idx]]
    lo <- quantile(m$df[[idx]], 0.02, na.rm = TRUE)
    hi <- quantile(m$df[[idx]], 0.98, na.rm = TRUE)
    pred_grid(idx, m$df, m$mod, lo, hi) %>% mutate(label = label_map[idx])
  }))

  fig3 <- ggplot(pp, aes(x = x, y = fit)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.14, fill = jacc_cols["blue"]) +
    geom_line(linewidth = 0.9, color = jacc_cols["navy"]) +
    facet_wrap(~ label, scales = "free_x", nrow = 3) +
    labs(title = bquote(bold("Nonlinear Association: Diastolic Indices and Peak" ~ VO[2])),
         x = NULL,
         y = bquote("Predicted Peak" ~ VO[2] ~ "(% Predicted, FRIEND 2.0)")) +
    theme_jacc()

  fig3
  save_jacc(fig3, "../2_Output/Figure3_Nonlinear_DiastolicIndices_FRIEND2.pdf", w = 6.9, h = 6.5)
}
```

## Secondary outcome: VE/VCO$_2$ slope

```{r cross_sectional_vevco2, fig.width=6.9, fig.height=6.5}
vevco2_models <- list()
vevco2_lrt <- list()

for (idx in diastolic_indices) {
  df_idx <- baseline_df %>%
    filter(!is.na(VeVco2_slope), !is.na(age), !is.na(Sex),
           !is.na(LBMI), !is.na(.data[[idx]]))
  if (nrow(df_idx) < 20) next

  base <- lm(VeVco2_slope ~ age + Sex + LBMI, data = df_idx)
  mod  <- lm(as.formula(paste0("VeVco2_slope ~ age + Sex + LBMI + ns(", idx, ", df=3)")),
             data = df_idx)
  lrt  <- anova(base, mod)

  vevco2_models[[idx]] <- list(df = df_idx, base = base, mod = mod, lrt = lrt)
  vevco2_lrt[[idx]] <- tibble(
    index = idx, label = label_map[idx], n = nrow(df_idx),
    delta_AIC = AIC(mod) - AIC(base),
    F_stat = lrt$F[2], p_value = lrt$`Pr(>F)`[2]
  )
}

vevco2_lrt_table <- bind_rows(vevco2_lrt) %>% arrange(p_value)
write.csv(vevco2_lrt_table, "../2_Output/Stats_CrossSectional_LRT_VeVco2.csv", row.names = FALSE)

if (length(vevco2_models) > 0) {
  pp_vevco2 <- bind_rows(lapply(names(vevco2_models), function(idx) {
    m <- vevco2_models[[idx]]
    lo <- quantile(m$df[[idx]], 0.02, na.rm = TRUE)
    hi <- quantile(m$df[[idx]], 0.98, na.rm = TRUE)
    pred_grid(idx, m$df, m$mod, lo, hi) %>% mutate(label = label_map[idx])
  }))

  fig_vevco2 <- ggplot(pp_vevco2, aes(x = x, y = fit)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, fill = jacc_cols["orange"]) +
    geom_line(linewidth = 0.9, color = jacc_cols["red"]) +
    facet_wrap(~ label, scales = "free_x", nrow = 3) +
    labs(title = bquote(bold("Nonlinear Association: Diastolic Indices and" ~ VE/VCO[2] ~ "Slope")),
         x = NULL,
         y = bquote(VE/VCO[2] ~ "Slope")) +
    theme_jacc()

  fig_vevco2
  save_jacc(fig_vevco2, "../2_Output/FigureS_Nonlinear_DiastolicIndices_VeVco2.pdf", w = 6.9, h = 6.5)
}
```

## Sensitivity: Echo alignment window

```{r sensitivity_window}
run_window <- function(w, idx, outcome_var = "VO2_FRIEND2_PP") {
  tmp <- cpx_aligned %>%
    left_join(echo_all, by = "MRN", relationship = "many-to-many") %>%
    mutate(delta_days = as.numeric(cpx_test_date - echo_date)) %>%
    filter(!is.na(delta_days), delta_days >= 0, delta_days <= w) %>%
    group_by(ID, CPX.sequential, cpx_test_date) %>%
    slice_min(order_by = delta_days, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(across(any_of(c(idx, "age", "LBMI", outcome_var)), ~ as.numeric(as.character(.))),
           Sex = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")))

  base <- tmp %>%
    group_by(ID) %>% arrange(days_from_baseline) %>% slice(1) %>% ungroup() %>%
    filter(!is.na(.data[[outcome_var]]), !is.na(.data[[idx]]), !is.na(age), !is.na(Sex), !is.na(LBMI))

  if (nrow(base) < 20) return(tibble(window_days = w, index = idx, n = nrow(base),
                                      delta_AIC = NA_real_, p_value = NA_real_))

  fit_base <- lm(as.formula(paste0(outcome_var, " ~ age + Sex + LBMI")), data = base)
  fit_idx  <- lm(as.formula(paste0(outcome_var, " ~ age + Sex + LBMI + ns(", idx, ", df=3)")), data = base)
  lrt <- anova(fit_base, fit_idx)

  tibble(window_days = w, index = idx, n = nrow(base),
         delta_AIC = AIC(fit_idx) - AIC(fit_base), p_value = lrt$`Pr(>F)`[2])
}

sens <- bind_rows(lapply(diastolic_indices, function(idx) {
  bind_rows(run_window(180, idx), run_window(365, idx), run_window(730, idx))
}))

write.csv(sens, "../2_Output/Sensitivity_EchoWindow_FRIEND2.csv", row.names = FALSE)
```

# Longitudinal GAMM Analysis

We employ a Generalized Additive Mixed Model (GAMM) via `mgcv::bam()` to model the nonlinear trajectory of Peak VO$_2$ over time, testing whether the trajectory diverges as a function of baseline diastolic severity via a tensor product interaction.

```{r gamm_analysis, fig.width=6.9, fig.height=5.5}
# Derive baseline DD composite Z-score from GAMLSS (if available) or standardized E/e'
if ("e_e_ave" %in% names(gamlss_results)) {
  # Use SEAMLSS Z-score from E/e' model
  z_source <- gamlss_results[["e_e_ave"]]$data %>%
    select(ID, dd_zscore = seamlss_z)
  long_df <- long_df %>%
    left_join(z_source, by = "ID")
} else {
  # Fallback: standardized E/e'
  long_df <- long_df %>%
    mutate(dd_zscore = scale(as.numeric(as.character(e_e_ave)))[, 1])
}

# Filter to patients with DD Z-score and sufficient follow-up
gamm_df <- long_df %>%
  filter(!is.na(VO2_FRIEND2_PP), !is.na(dd_zscore), !is.na(time_yrs),
         !is.na(age), !is.na(LBMI), is.finite(time_yrs)) %>%
  mutate(ID_fac = factor(ID))

cat(sprintf("GAMM cohort: %d observations, %d patients\n",
            nrow(gamm_df), n_distinct(gamm_df$ID)))

if (nrow(gamm_df) >= 50 && n_distinct(gamm_df$ID) >= 20) {
  # Fit GAMM with tensor product interaction
  gamm_fit <- tryCatch({
    bam(VO2_FRIEND2_PP ~ s(time_yrs, bs = "cr", k = 5) +
          s(dd_zscore, bs = "cr", k = 5) +
          ti(time_yrs, dd_zscore, k = c(4, 4)) +
          age + Sex + LBMI +
          s(ID_fac, bs = "re"),
        data = gamm_df, method = "fREML", discrete = TRUE)
  }, error = function(e) {
    message("GAMM failed, fitting simpler model: ", e$message)
    bam(VO2_FRIEND2_PP ~ s(time_yrs, bs = "cr", k = 5) +
          dd_zscore * time_yrs +
          age + Sex + LBMI +
          s(ID_fac, bs = "re"),
        data = gamm_df, method = "fREML", discrete = TRUE)
  })

  # Comparison: linear mixed model
  lme_fit <- lmer(VO2_FRIEND2_PP ~ time_yrs * dd_zscore + age + Sex + LBMI + (1 | ID),
                  data = gamm_df, REML = FALSE)

  cat(sprintf("GAMM AIC: %.1f  |  LME AIC: %.1f\n", AIC(gamm_fit), AIC(lme_fit)))

  # Summary of smooth terms
  gamm_summary <- summary(gamm_fit)
  cat("\n**GAMM Smooth Terms:**\n")
  print(gamm_summary$s.table)

  write.csv(as.data.frame(gamm_summary$s.table),
            "../2_Output/Stats_GAMM_SmoothTerms.csv")
  write.csv(as.data.frame(gamm_summary$p.table),
            "../2_Output/Stats_GAMM_ParametricTerms.csv")

  # Prediction: trajectories stratified by DD Z-score tertiles
  z_tertiles <- quantile(gamm_df$dd_zscore, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)

  pred_gamm <- expand_grid(
    time_yrs  = seq(0, min(5, max(gamm_df$time_yrs, na.rm = TRUE)), by = 0.25),
    dd_zscore = z_tertiles
  ) %>%
    mutate(
      level = factor(dd_zscore, labels = c("Q1 (less impaired)", "Median", "Q3 (more impaired)")),
      age  = median(gamm_df$age, na.rm = TRUE),
      Sex  = factor(names(which.max(table(gamm_df$Sex))), levels = levels(gamm_df$Sex)),
      LBMI = median(gamm_df$LBMI, na.rm = TRUE),
      ID_fac = gamm_df$ID_fac[1]
    )

  pred_gamm$fit <- predict(gamm_fit, newdata = pred_gamm, exclude = "s(ID_fac)", type = "response")

  fig4 <- ggplot(pred_gamm, aes(x = time_yrs, y = fit, color = level)) +
    geom_line(linewidth = 1.1) +
    scale_color_manual(values = c(jacc_cols["blue"], jacc_cols["navy"], jacc_cols["red"])) +
    labs(title = bquote(bold("Longitudinal Peak" ~ VO[2] ~ "Trajectories by Baseline DD Severity")),
         subtitle = "GAMM with tensor product interaction: time \u00D7 DD Z-score",
         x = "Years from Baseline CPET",
         y = bquote("Predicted Peak" ~ VO[2] ~ "(FRIEND 2.0 % Pred)"),
         color = "Baseline DD Z-score") +
    theme_jacc() +
    theme(legend.position = "right")

  fig4
  save_jacc(fig4, "../2_Output/Figure4_GAMM_Trajectories.pdf", w = 6.9, h = 5.0)
}
```

# Heart Failure Outcome Analysis

Time-to-event analysis tests whether ASE 2025 DD grade and continuous DD Z-scores predict a composite heart failure endpoint (acute HF, chronic HF, transplant, or death).

```{r hf_outcomes, fig.width=6.9, fig.height=5.0, results='asis'}
surv_df <- baseline_df %>%
  filter(!is.na(hf_composite), !is.na(hf_composite_yrs) | hf_composite == 0) %>%
  mutate(
    # For censored patients, use time from CPX to last encounter
    follow_up_yrs = ifelse(
      hf_composite == 1 & !is.na(hf_composite_yrs),
      hf_composite_yrs,
      as.numeric(as.Date(last_enc_date) - cpx_test_date) / 365.25
    ),
    follow_up_yrs = ifelse(is.na(follow_up_yrs) | follow_up_yrs <= 0, 0.01, follow_up_yrs)
  ) %>%
  filter(!is.na(follow_up_yrs), follow_up_yrs > 0)

cat(sprintf("Survival cohort: %d patients, %d events\n",
            nrow(surv_df), sum(surv_df$hf_composite == 1, na.rm = TRUE)))

if (nrow(surv_df) >= 30 && sum(surv_df$hf_composite == 1, na.rm = TRUE) >= 5) {

  # Kaplan-Meier by DD grade
  surv_grade <- surv_df %>% filter(dd_grade != "Indeterminate", !is.na(dd_grade))

  if (nrow(surv_grade) >= 20) {
    km_fit <- survfit(Surv(follow_up_yrs, hf_composite) ~ dd_grade, data = surv_grade)

    # KM plot
    km_data <- broom::tidy(km_fit) %>%
      mutate(strata = str_remove(strata, "dd_grade="))

    fig5 <- ggplot(km_data, aes(x = time, y = estimate, color = strata)) +
      geom_step(linewidth = 0.9) +
      geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = strata),
                  alpha = 0.1, stat = "identity") +
      scale_color_manual(values = c("Normal" = jacc_cols["blue"],
                                     "Grade I" = jacc_cols["green"],
                                     "Grade II" = jacc_cols["orange"],
                                     "Grade III" = jacc_cols["red"])) +
      scale_fill_manual(values = c("Normal" = jacc_cols["blue"],
                                    "Grade I" = jacc_cols["green"],
                                    "Grade II" = jacc_cols["orange"],
                                    "Grade III" = jacc_cols["red"])) +
      scale_y_continuous(limits = c(0, 1), labels = percent) +
      labs(title = "Composite HF Endpoint by ASE 2025 DD Grade",
           x = "Years from Baseline CPET",
           y = "Event-Free Survival",
           color = "DD Grade", fill = "DD Grade") +
      theme_jacc() +
      theme(legend.position = "right")

    fig5
    save_jacc(fig5, "../2_Output/Figure5_KM_DDGrade.pdf", w = 6.9, h = 5.0)

    # Log-rank test
    lr_test <- survdiff(Surv(follow_up_yrs, hf_composite) ~ dd_grade, data = surv_grade)
    cat(sprintf("\n**Log-rank test p-value: %.4f**\n\n", 1 - pchisq(lr_test$chisq, df = length(lr_test$n) - 1)))
  }

  # Cox proportional hazards: DD grade
  cox_grade <- tryCatch({
    coxph(Surv(follow_up_yrs, hf_composite) ~ dd_grade + age + Sex + LBMI,
          data = surv_grade)
  }, error = function(e) NULL)

  if (!is.null(cox_grade)) {
    cat("\n**Cox Model: DD Grade**\n")
    tidy(cox_grade, exponentiate = TRUE, conf.int = TRUE) %>%
      mutate(across(where(is.numeric), ~ round(., 3))) %>%
      flextable() %>%
      bold(part = "header") %>%
      theme_booktabs() %>%
      autofit() %>%
      set_caption("Table 3. Cox Proportional Hazards: Composite HF Endpoint") %>%
      print()

    write.csv(tidy(cox_grade, exponentiate = TRUE, conf.int = TRUE),
              "../2_Output/Table3_Cox_DDGrade.csv", row.names = FALSE)

    # PH assumption check
    ph_check <- cox.zph(cox_grade)
    cat(sprintf("\n**Proportional hazards global test p = %.3f**\n", ph_check$table["GLOBAL", "p"]))
  }

  # Cox: continuous DD Z-score (from GAMLSS if available)
  if ("e_e_ave" %in% names(gamlss_results)) {
    z_src <- gamlss_results[["e_e_ave"]]$data %>% select(ID, dd_zscore = seamlss_z)
    surv_z <- surv_df %>% left_join(z_src, by = "ID") %>% filter(!is.na(dd_zscore))

    if (nrow(surv_z) >= 20 && sum(surv_z$hf_composite == 1, na.rm = TRUE) >= 5) {
      cox_z <- coxph(Surv(follow_up_yrs, hf_composite) ~ dd_zscore + age + Sex + LBMI,
                     data = surv_z)
      cat("\n**Cox Model: Continuous DD Z-Score**\n")
      tidy(cox_z, exponentiate = TRUE, conf.int = TRUE) %>%
        mutate(across(where(is.numeric), ~ round(., 3))) %>%
        flextable() %>% bold(part = "header") %>% theme_booktabs() %>% autofit() %>% print()

      write.csv(tidy(cox_z, exponentiate = TRUE, conf.int = TRUE),
                "../2_Output/Table3_Cox_DDZscore.csv", row.names = FALSE)
    }
  }
} else {
  cat("Insufficient events for survival analysis.\n")
}
```

# Patient Characteristics

## Table 1

```{r table1, results='asis'}
table1_data <- baseline_df %>%
  mutate(
    VeVco2_slope = as.numeric(as.character(VeVco2_slope)),
    HCM_Phenotype = as.factor(HCM_Phenotype)
  ) %>%
  select(dd_grade, age, Sex, BMI, BSA, LBMI, HCM_Phenotype,
         pk.RER, VO2_FRIEND2_PP, VO2_WASSERMAN_PP, VeVco2_slope, HRR,
         e_e_ave, la_vol_index, E_vel, tr_max_vel, ef_modsp4)

# Overall
table1_overall <- table1_data %>%
  select(-dd_grade) %>%
  tbl_summary(
    label = list(
      age              ~ "Age, years",
      Sex              ~ "Sex",
      BMI              ~ "Body Mass Index, kg/m\u00B2",
      BSA              ~ "Body Surface Area, m\u00B2",
      LBMI             ~ "Lean Body Mass Index, kg/m\u00B2",
      HCM_Phenotype    ~ "HCM Phenotype",
      pk.RER           ~ "Peak RER",
      VO2_FRIEND2_PP   ~ "Peak VO\u2082 (FRIEND 2.0 %pred)",
      VO2_WASSERMAN_PP ~ "Peak VO\u2082 (Wasserman %pred)",
      VeVco2_slope     ~ "VE/VCO\u2082 Slope",
      HRR              ~ "Heart Rate Recovery (1 min)",
      e_e_ave          ~ "E/e' (average)",
      la_vol_index     ~ "LAVI, mL/m\u00B2",
      E_vel            ~ "Mitral E Velocity, cm/s",
      tr_max_vel       ~ "TR Vmax, m/s",
      ef_modsp4        ~ "LVEF, %"
    ),
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 1,
    missing = "no"
  ) %>%
  bold_labels() %>%
  modify_caption("**Table 1. Baseline Patient Characteristics**")

table1_overall

table1_overall %>%
  as_flex_table() %>%
  save_as_docx(path = "../2_Output/Table1_Manuscript.docx")
```

## Figure 1: Cohort Characteristics

```{r figure1, fig.width=6.9, fig.height=7.5}
accent  <- jacc_cols["blue"]
gray_f  <- "grey78"

# Panel A: Age
p_age <- ggplot(table1_data, aes(x = age)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 5, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = accent, linewidth = 0.8) +
  labs(title = "Age", x = "Years", y = "Density") +
  theme_jacc()

# Panel B: BMI
p_bmi <- ggplot(table1_data, aes(x = BMI)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 2, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = accent, linewidth = 0.8) +
  labs(title = "BMI", x = bquote("kg/" * m^2), y = "Density") +
  theme_jacc()

# Panel C: LBMI
p_lbmi <- ggplot(table1_data, aes(x = LBMI)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = accent, linewidth = 0.8) +
  labs(title = "LBMI", x = bquote("kg/" * m^2), y = "Density") +
  theme_jacc()

# Panel D: Peak VO2 comparison
vo2_long <- table1_data %>%
  select(VO2_FRIEND2_PP, VO2_WASSERMAN_PP) %>%
  pivot_longer(everything(), names_to = "Equation", values_to = "Value") %>%
  mutate(Equation = recode(Equation,
                            "VO2_FRIEND2_PP" = "FRIEND 2.0",
                            "VO2_WASSERMAN_PP" = "Wasserman"))

p_vo2 <- ggplot(vo2_long, aes(x = Value, fill = Equation)) +
  geom_histogram(aes(y = after_stat(density)), bins = 25, alpha = 0.35, position = "identity", color = "white") +
  geom_density(aes(color = Equation), linewidth = 0.6) +
  scale_fill_manual(values = c("FRIEND 2.0" = jacc_cols["blue"], "Wasserman" = jacc_cols["gray"])) +
  scale_color_manual(values = c("FRIEND 2.0" = jacc_cols["navy"], "Wasserman" = jacc_cols["gray"])) +
  scale_x_continuous(limits = c(0, 200)) +
  labs(title = bquote(bold("Peak" ~ VO[2] ~ "Comparison")),
       x = bquote("% Predicted" ~ VO[2]), y = "Density") +
  theme_jacc()

# Panel E: VE/VCO2
p_vevco2 <- ggplot(table1_data, aes(x = VeVco2_slope)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = accent, linewidth = 0.8) +
  labs(title = bquote(bold(VE/VCO[2] ~ "Slope")), x = bquote(VE/VCO[2]), y = "Density") +
  theme_jacc()

# Panel F-I: Echo distributions
p_ee <- ggplot(table1_data, aes(x = e_e_ave)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = jacc_cols["red"], linewidth = 0.8) +
  geom_vline(xintercept = 14, linetype = "dashed", color = jacc_cols["red"], linewidth = 0.4) +
  labs(title = "E/e' (average)", x = "E/e'", y = "Density") +
  theme_jacc()

p_lavi <- ggplot(table1_data, aes(x = la_vol_index)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = jacc_cols["red"], linewidth = 0.8) +
  geom_vline(xintercept = 34, linetype = "dashed", color = jacc_cols["red"], linewidth = 0.4) +
  labs(title = "LAVI", x = bquote("mL/" * m^2), y = "Density") +
  theme_jacc()

p_trv <- ggplot(table1_data, aes(x = tr_max_vel)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = jacc_cols["red"], linewidth = 0.8) +
  geom_vline(xintercept = 2.8, linetype = "dashed", color = jacc_cols["red"], linewidth = 0.4) +
  labs(title = "TR Vmax", x = "m/s", y = "Density") +
  theme_jacc()

p_ef <- ggplot(table1_data, aes(x = ef_modsp4)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = accent, linewidth = 0.8) +
  labs(title = "LVEF", x = "%", y = "Density") +
  theme_jacc()

fig1 <- (p_age | p_bmi | p_lbmi) /
        (p_vo2 | p_vevco2 | p_ee) /
        (p_lavi | p_trv | p_ef) +
  plot_annotation(
    tag_levels = "A",
    title = "Baseline Cohort Biometrics, Functional Capacity, and Echocardiography"
  ) &
  theme(plot.tag = element_text(face = "bold", size = 10))

fig1
save_jacc(fig1, "../2_Output/Figure1_Cohort_Characteristics.pdf", w = 6.9, h = 7.5)
```

## Figure 2: Longitudinal Analysis

```{r figure2, fig.width=6.9, fig.height=5.5}
# Annualized slopes
CPX_slopes <- cpx_echo %>%
  mutate(VO2_FRIEND2_PP = as.numeric(as.character(VO2_FRIEND2_PP)),
         VeVco2_slope = as.numeric(as.character(VeVco2_slope))) %>%
  group_by(ID) %>%
  arrange(days_from_baseline) %>%
  summarize(
    follow_up_years    = max(days_from_baseline) / 365.25,
    annual_delta_vo2   = (last(VO2_FRIEND2_PP) - first(VO2_FRIEND2_PP)) / (max(days_from_baseline) / 365.25),
    baseline_vo2       = first(VO2_FRIEND2_PP),
    baseline_vevco2    = first(VeVco2_slope),
    num_tests          = n(),
    .groups = "drop"
  ) %>%
  filter(num_tests > 1, is.finite(annual_delta_vo2))

# Panel A: Spaghetti plot
p_spaghetti <- ggplot(cpx_echo %>% mutate(VO2_FRIEND2_PP = as.numeric(as.character(VO2_FRIEND2_PP))),
                       aes(x = days_from_baseline / 365.25, y = VO2_FRIEND2_PP)) +
  geom_line(aes(group = ID), alpha = 0.15, color = jacc_cols["navy"]) +
  geom_point(alpha = 0.25, size = 0.8, color = jacc_cols["blue"]) +
  geom_smooth(method = "lm", color = jacc_cols["red"], linewidth = 1, se = TRUE,
              fill = jacc_cols["red"], alpha = 0.1) +
  labs(title = bquote(bold("Peak" ~ VO[2] ~ "Trajectories")),
       x = "Years from Baseline", y = bquote("Peak" ~ VO[2] ~ "(% Pred)")) +
  theme_jacc()

# Panel B: Swimmer plot
p_swimmer <- ggplot(cpx_echo, aes(x = days_from_baseline / 365.25, y = as.factor(ID))) +
  geom_line(aes(group = ID), color = jacc_cols["navy"], linewidth = 0.3) +
  geom_point(aes(color = as.numeric(as.character(VO2_FRIEND2_PP))), size = 1) +
  scale_color_viridis_c(option = "magma", name = "% Pred") +
  labs(title = "Patient Timeline", x = "Years Since Baseline", y = NULL) +
  theme_jacc() + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

# Panel C: Correlation heatmap
corr_data <- CPX_slopes %>%
  left_join(table1_data %>% select(any_of(c("ID", "age", "BMI", "LBMI"))),
            by = character()) %>%  # will use table1_data from baseline
  select(any_of(c("baseline_vo2", "baseline_vevco2", "annual_delta_vo2", "follow_up_years")))

# Use baseline_df for correlation
corr_src <- baseline_df %>%
  left_join(CPX_slopes %>% select(ID, annual_delta_vo2, follow_up_years), by = "ID") %>%
  select(age, BMI, LBMI,
         `Baseline VO2%` = VO2_FRIEND2_PP,
         `VE/VCO2` = VeVco2_slope,
         `Annual dVO2%` = annual_delta_vo2,
         `Follow-up (Yrs)` = follow_up_years) %>%
  mutate(across(everything(), ~ as.numeric(as.character(.)))) %>%
  drop_na()

if (nrow(corr_src) >= 10) {
  corr_matrix <- round(cor(corr_src, use = "pairwise.complete.obs"), 2)
  p_corr <- ggcorrplot(corr_matrix, hc.order = TRUE, type = "lower", lab = TRUE,
                        lab_size = 2.5,
                        colors = c(jacc_cols["red"], "white", jacc_cols["green"]),
                        title = "Parameter Correlations") +
    theme_jacc() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
          axis.text.y = element_text(size = 7))
} else {
  p_corr <- ggplot() + annotate("text", x = 0.5, y = 0.5, label = "Insufficient data") + theme_void()
}

fig2 <- (p_spaghetti | p_swimmer | p_corr) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "bold", size = 11))

fig2
save_jacc(fig2, "../2_Output/Figure2_Longitudinal_Analysis.pdf", w = 6.9, h = 5.5)
```

# Supplemental Analyses

## Age-residualized diastolic indices

```{r residualize, fig.width=6.9, fig.height=4.5}
baseline_df_resid <- baseline_df
for (idx in diastolic_indices) {
  baseline_df_resid <- residualize_on_age(baseline_df_resid, idx)
}

resid_summary <- bind_rows(lapply(diastolic_indices, function(idx) {
  resid_col <- paste0(idx, "_resid")
  tibble(
    index    = idx,
    label    = label_map[idx],
    n_raw    = sum(!is.na(baseline_df[[idx]])),
    n_resid  = sum(!is.na(baseline_df_resid[[resid_col]])),
    sd_resid = round(sd(baseline_df_resid[[resid_col]], na.rm = TRUE), 3)
  )
}))

resid_summary %>%
  flextable() %>%
  bold(part = "header") %>%
  theme_booktabs() %>%
  autofit() %>%
  set_caption("Table S. Age-Residualized Diastolic Indices Summary")

# Diagnostic boxplot
resid_long <- baseline_df_resid %>%
  select(all_of(paste0(diastolic_indices, "_resid"))) %>%
  pivot_longer(everything(), names_to = "index", values_to = "resid") %>%
  mutate(index = str_replace(index, "_resid$", ""),
         label = label_map[index])

p_resid <- ggplot(resid_long, aes(x = label, y = resid)) +
  geom_boxplot(fill = "grey85", color = "black", outlier.size = 0.5, linewidth = 0.3) +
  coord_flip() +
  labs(title = "Age-Residualized Diastolic Indices", x = NULL, y = "Residual") +
  theme_jacc()

p_resid
save_jacc(p_resid, "../2_Output/FigureS_AgeResidualized_DiastolicIndices.pdf", w = 6.9, h = 4.5)
```

## Export missing echo audit

```{r export_missing_echo}
echo_vars_present <- intersect(
  c("e_e_ave", "e_e_lat", "e_e_med", "med_peak_e_vel", "ef_modsp4",
    "la_vol", "la_vol_index", "lat_peak_e_vel", "mv_a_point", "mv_a_dur",
    "mv_dec_time", "mv_e_a", "pulm_dias_vel", "tr_max_vel"),
  colnames(cpx_echo)
)

missing_echo <- cpx_aligned %>%
  anti_join(cpx_echo %>% select(ID, cpx_test_date) %>% distinct(),
            by = c("ID", "cpx_test_date")) %>%
  arrange(MRN, cpx_test_date) %>%
  select(ID, MRN, CPX.sequential, cpx_test_date, days_from_baseline, age, Sex, BMI, LBMI,
         VO2_FRIEND2_PP, VeVco2_slope)

if (nrow(missing_echo) > 0) {
  openxlsx::write.xlsx(
    list(Missing_Echo_CPX = missing_echo,
         Summary = tibble(n_missing = nrow(missing_echo),
                          n_patients = n_distinct(missing_echo$ID))),
    file = "../2_Output/Serial_CPX_missing_resting_echo.xlsx",
    overwrite = TRUE
  )
  cat(sprintf("Exported %d CPX tests without echo linkage (%d patients)\n",
              nrow(missing_echo), n_distinct(missing_echo$ID)))
}
```

```{r session_info}
cat("\n\n---\n**Session Info:**\n")
sessionInfo()
```
