---
title: "Ergocycle-Based Cardiopulmonary Exercise Testing and Clinical Outcomes"
author: "Mark E. Pepin, MD, PhD, MS, FESC"
date: "2026-03-23"
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
    code-tools: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
  results: hide
---

This analysis examines ergocycle-based cardiopulmonary exercise tests (CPX) captured in the study workbook and links those tests to the available clinical outcomes table. All cycle-based protocols are retained irrespective of diagnostic subgroup or coexisting conditions. The analytic emphasis is descriptive and exploratory, focusing on cohort assembly, baseline CPX phenotype, and observed post-CPX clinical events.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 5,
  dpi = 300
)

library(openxlsx)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(gtsummary)
library(flextable)
library(patchwork)
library(pROC)
library(survival)

output_dir <- "../2_Output/Cycle"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
output_file <- function(filename) file.path(output_dir, filename)
km_max_years <- 10

group_colors <- c(
  "No HF/PHTN" = "#4C78A8",
  "HF" = "#D95F02",
  "PHTN" = "#B279A2"
)

group_display <- c(
  "No HF/PHTN" = "CON",
  "HF" = "CHF",
  "PHTN" = "PH"
)

group_display_colors <- c(
  "CON" = "#4C78A8",
  "CHF" = "#E15759",
  "PH" = "#59A14F"
)

theme_publication <- function() {
  theme_bw(base_size = 12) +
    theme(
      text = element_text(color = "grey10"),
      axis.line = element_line(color = "grey10", linewidth = 0.5),
      axis.ticks = element_line(color = "grey10", linewidth = 0.4),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank(),
      panel.border = element_rect(color = "grey10", linewidth = 0.5),
      plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
      axis.title = element_text(face = "bold"),
      strip.background = element_rect(fill = "grey94", color = NA),
      strip.text = element_text(face = "bold"),
      legend.position = "top",
      legend.title = element_blank()
    )
}
```

## Cohort assembly

Cycle studies were identified using the structured treadmill-versus-cycle field together with explicit protocol labels indicating bike, cycle, or ergometer testing. Obvious walking or running entries were excluded when they conflicted with the cycling designation. A stable patient identifier was generated from MRN for patient-level summaries. No filtering was applied on the basis of HCM diagnosis or noncardiac comorbidity.

```{r cohort}
CPX_raw <- read.xlsx("../1_Input/CPX_input.xlsx", sheet = "CPX", detectDates = TRUE)
Outcomes_raw <- read.xlsx("../1_Input/CPX_input.xlsx", sheet = "Outcomes_2025", detectDates = TRUE)
Comorbidities_raw <- read.xlsx("../1_Input/CPX_input.xlsx", sheet = "co-morbidities", detectDates = TRUE)

cycling_pattern <- "cycl|bike|erg|supine|upright|ice"
noncycling_modes <- c("walking", "running", "vo2 study")

CPX_cycle <- CPX_raw %>%
  mutate(
    mode_clean = tolower(trimws(dplyr::coalesce(Mode, ""))),
    protocol_clean = tolower(trimws(dplyr::coalesce(Protocol, ""))),
    cycle_flag = suppressWarnings(as.numeric(`Treadmill.(0).or.cycle.(1)`)),
    is_cycling_test =
      ((cycle_flag == 1) & !(mode_clean %in% noncycling_modes)) |
      grepl(cycling_pattern, mode_clean) |
      grepl(cycling_pattern, protocol_clean),
    Exercise_mode = if_else(
      nzchar(mode_clean),
      Mode,
      if_else(cycle_flag == 1, "Cycling", NA_character_)
    ),
    Cycling_form = case_when(
      grepl("supine", protocol_clean) ~ "Supine bike",
      grepl("upright", protocol_clean) ~ "Upright bike",
      grepl("(^|[^a-z])ice([^a-z]|$)", protocol_clean) ~ "ICE/research bike",
      grepl("manual", protocol_clean) ~ "Manual bike",
      is_cycling_test ~ "Other/unspecified cycling",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(is_cycling_test) %>%
  mutate(MRN = as.character(MRN)) %>%
  group_by(MRN) %>%
  mutate(ID = cur_group_id()) %>%
  ungroup() %>%
  mutate(
    cpx_test_date = as.Date(cpx_test_date),
    BSA = sqrt((`height.(cm)` * `Weight.(kg)`) / 3600),
    BMI = `Weight.(kg)` / ((`height.(cm)` / 100) ^ 2),
    LBM = LBM.NHANES.no.race,
    LBMI = LBM / ((`height.(cm)` / 100) ^ 2)
  ) %>%
  select(
    ID, MRN, cpx_test_date, age,
    Sex = `Sex:.M(0)/F(1)`,
    Exercise_mode, Cycling_form, Protocol, CPX.sequential,
    BMI, BSA, LBM, LBMI,
    pk.RER,
    VO2_kg = `VO2.(ml/min/kg)`,
    VO2_FRIEND_PP = `FRIEND1.%.predicted.VO2`,
    VO2_FRIEND2_PP = `FRIEND2.%predicted.VO2`,
    VO2_WASSERMAN_PP = `Wasserman.%predicted.VO2.(2005)`,
    HRmax_PP = `%predicted.HR.FRIEND`,
    HRR = `HR.recovery.(1min)`,
    VeVco2_slope = `ve/vco2.slope`,
    OUES, `Bike.Watts`, Arrhythmia, Notes
  ) %>%
  filter(
    is.na(VO2_FRIEND2_PP) | dplyr::between(VO2_FRIEND2_PP, 0, 200),
    is.na(VO2_FRIEND_PP) | dplyr::between(VO2_FRIEND_PP, 0, 200),
    is.na(VO2_WASSERMAN_PP) | dplyr::between(VO2_WASSERMAN_PP, 0, 200)
  )

Outcomes_clean <- Outcomes_raw %>%
  transmute(
    MRN = as.character(MRN),
    cpx_test_date = as.Date(cpx_test_date, origin = "1899-12-30"),
    last_enc_date = as.Date(last_enc_date, origin = "1899-12-30"),
    death = as.numeric(death),
    death_yrs = as.numeric(death_yrs),
    post_acute_heart_failure = as.numeric(post_acute_heart_failure),
    post_acute_heart_failure_yrs = as.numeric(post_acute_heart_failure_yrs),
    post_cardiac_arrest = as.numeric(post_cardiac_arrest),
    post_cardiac_arrest_yrs = as.numeric(post_cardiac_arrest_yrs),
    post_heart_transplant = as.numeric(post_heart_transplant),
    post_heart_transplant_yrs = as.numeric(post_heart_transplant_yrs),
    post_ventricular_tachycardia = as.numeric(post_ventricular_tachycardia),
    post_ventricular_tachycardia_yrs = as.numeric(post_ventricular_tachycardia_yrs),
    post_ventricular_fib_flut_tachy = as.numeric(post_ventricular_fib_flut_tachy),
    post_ventricular_fib_flut_tachy_yrs = as.numeric(post_ventricular_fib_flut_tachy_yrs),
    post_defibrillator = as.numeric(post_defibrillator),
    post_defibrillator_yrs = as.numeric(post_defibrillator_yrs),
    post_afib_flut = as.numeric(post_afib_flut),
    post_afib_flut_yrs = as.numeric(post_afib_flut_yrs),
    post_septal_reduction_surgery = as.numeric(post_septal_reduction_surgery),
    post_septal_reduction_surgery_yrs = as.numeric(post_septal_reduction_surgery_yrs)
  )

Comorbidities_clean <- Comorbidities_raw %>%
  transmute(
    MRN = as.character(MRN),
    cpx_test_date = as.Date(cpx_test_date),
    hf_dx = as.numeric(hf_pre_test),
    phtn_dx = as.numeric(
      pmax(
        dplyr::coalesce(as.numeric(pah_pre_test), 0),
        dplyr::coalesce(as.numeric(pahtn_pre_test), 0),
        dplyr::coalesce(as.numeric(pulm_htn_pre_test), 0),
        na.rm = TRUE
      )
    )
  )

CPX_cycle <- CPX_cycle %>%
  left_join(Comorbidities_clean, by = c("MRN", "cpx_test_date")) %>%
  left_join(Outcomes_clean, by = c("MRN", "cpx_test_date")) %>%
  mutate(
    Disease_group = case_when(
      phtn_dx == 1 ~ "PHTN",
      hf_dx == 1 ~ "HF",
      TRUE ~ "No HF/PHTN"
    ),
    Disease_group = factor(Disease_group, levels = c("No HF/PHTN", "HF", "PHTN"))
  )

baseline_cohort <- CPX_cycle %>%
  group_by(ID) %>%
  arrange(cpx_test_date, .by_group = TRUE) %>%
  slice(1) %>%
  ungroup()

cohort_flow <- dplyr::bind_rows(
  CPX_cycle %>%
    count(Disease_group, name = "n_tests") %>%
    mutate(stage = "All cycle-based CPX tests"),
  CPX_cycle %>%
    filter(!is.na(death)) %>%
    count(Disease_group, name = "n_tests") %>%
    mutate(stage = "Cycle-based CPX tests linked to outcomes"),
  baseline_cohort %>%
    count(Disease_group, name = "n_tests") %>%
    mutate(stage = "Baseline cohort with phenotype assignment")
  ) %>%
  mutate(
    stage = factor(
      stage,
      levels = c(
        "Baseline cohort with phenotype assignment",
        "Cycle-based CPX tests linked to outcomes",
        "All cycle-based CPX tests"
      )
    ),
    Disease_group = factor(Disease_group, levels = c("No HF/PHTN", "HF", "PHTN")),
    group_label = factor(group_display[as.character(Disease_group)], levels = c("CON", "CHF", "PH"))
  ) %>%
  group_by(stage) %>%
  mutate(total_tests = sum(n_tests)) %>%
  ungroup()
```

## Exported linked test table

The final Excel workbook was structured for both descriptive review and downstream prediction modeling. It includes a test-level sheet for all linked ergocycle CPX studies, a baseline patient-level sheet for one-row-per-patient modeling, and a compact data dictionary.

```{r export_linked_tests}
linked_cycle_tests <- CPX_cycle %>%
  mutate(
    Sex = case_when(
      Sex == 0 ~ "Male",
      Sex == 1 ~ "Female",
      TRUE ~ NA_character_
    ),
    ventricular_arrhythmia_composite = pmax(
      dplyr::coalesce(post_ventricular_tachycardia, 0),
      dplyr::coalesce(post_ventricular_fib_flut_tachy, 0),
      na.rm = TRUE
    ),
    ventricular_arrhythmia_yrs = pmin(
      if_else(post_ventricular_tachycardia == 1, post_ventricular_tachycardia_yrs, Inf),
      if_else(post_ventricular_fib_flut_tachy == 1, post_ventricular_fib_flut_tachy_yrs, Inf),
      na.rm = TRUE
    ),
    primary_composite = pmax(
      dplyr::coalesce(death, 0),
      dplyr::coalesce(post_cardiac_arrest, 0),
      dplyr::coalesce(post_heart_transplant, 0),
      na.rm = TRUE
    ),
    primary_composite_yrs = pmin(
      if_else(death == 1, death_yrs, Inf),
      if_else(post_cardiac_arrest == 1, post_cardiac_arrest_yrs, Inf),
      if_else(post_heart_transplant == 1, post_heart_transplant_yrs, Inf),
      na.rm = TRUE
    ),
    ventricular_arrhythmia_yrs = ifelse(is.infinite(ventricular_arrhythmia_yrs), NA_real_, ventricular_arrhythmia_yrs),
    primary_composite_yrs = ifelse(is.infinite(primary_composite_yrs), NA_real_, primary_composite_yrs),
    follow_up_yrs = as.numeric(last_enc_date - cpx_test_date) / 365.25
  ) %>%
  filter(!is.na(death)) %>%
  arrange(MRN, cpx_test_date, CPX.sequential) %>%
  mutate(
    across(
      c(
        death, post_acute_heart_failure, post_cardiac_arrest, post_heart_transplant,
        post_ventricular_tachycardia, post_ventricular_fib_flut_tachy, post_defibrillator,
        post_afib_flut, post_septal_reduction_surgery, primary_composite,
        ventricular_arrhythmia_composite, hf_dx, phtn_dx
      ),
      ~ case_when(.x == 1 ~ "Yes", .x == 0 ~ "No", TRUE ~ NA_character_)
    )
  ) %>%
  rename(
    Patient_ID = ID,
    Test_Date = cpx_test_date,
    Test_Sequence = CPX.sequential,
    Peak_RER = pk.RER,
    Peak_VO2_mL_kg_min = VO2_kg,
    Peak_VO2_FRIEND1_pct = VO2_FRIEND_PP,
    Peak_VO2_FRIEND2_pct = VO2_FRIEND2_PP,
    Peak_VO2_Wasserman2005_pct = VO2_WASSERMAN_PP,
    Peak_HR_pct_predicted = HRmax_PP,
    HR_Recovery_1min = HRR,
    VE_VCO2_Slope = VeVco2_slope,
    Bike_Watts = `Bike.Watts`,
    Last_Encounter_Date = last_enc_date,
    Death = death,
    Death_Years = death_yrs,
    Acute_Heart_Failure = post_acute_heart_failure,
    Acute_Heart_Failure_Years = post_acute_heart_failure_yrs,
    Cardiac_Arrest = post_cardiac_arrest,
    Cardiac_Arrest_Years = post_cardiac_arrest_yrs,
    Heart_Transplant = post_heart_transplant,
    Heart_Transplant_Years = post_heart_transplant_yrs,
    Ventricular_Tachycardia = post_ventricular_tachycardia,
    Ventricular_Tachycardia_Years = post_ventricular_tachycardia_yrs,
    Ventricular_Fib_Flutter_Tachy = post_ventricular_fib_flut_tachy,
    Ventricular_Fib_Flutter_Tachy_Years = post_ventricular_fib_flut_tachy_yrs,
    Defibrillator = post_defibrillator,
    Defibrillator_Years = post_defibrillator_yrs,
    AFib_Flutter = post_afib_flut,
    AFib_Flutter_Years = post_afib_flut_yrs,
    Septal_Reduction_Surgery = post_septal_reduction_surgery,
    Septal_Reduction_Surgery_Years = post_septal_reduction_surgery_yrs,
    Heart_Failure_PreTest = hf_dx,
    Pulmonary_Hypertension_PreTest = phtn_dx,
    Ventricular_Arrhythmia_Composite = ventricular_arrhythmia_composite,
    Ventricular_Arrhythmia_Composite_Years = ventricular_arrhythmia_yrs,
    Primary_Composite = primary_composite,
    Primary_Composite_Years = primary_composite_yrs,
    FollowUp_Years = follow_up_yrs
  ) %>%
  select(
    Patient_ID, MRN, Test_Date, Test_Sequence, Disease_group,
    age, Sex, Exercise_mode, Cycling_form, Protocol,
    BMI, BSA, LBM, LBMI, Peak_RER, Peak_VO2_mL_kg_min,
    Peak_VO2_FRIEND1_pct, Peak_VO2_FRIEND2_pct, Peak_VO2_Wasserman2005_pct,
    Peak_HR_pct_predicted, HR_Recovery_1min, VE_VCO2_Slope, OUES, Bike_Watts,
    Arrhythmia, Notes, Heart_Failure_PreTest, Pulmonary_Hypertension_PreTest,
    Last_Encounter_Date, FollowUp_Years,
    Primary_Composite, Primary_Composite_Years,
    Ventricular_Arrhythmia_Composite, Ventricular_Arrhythmia_Composite_Years,
    Death, Death_Years,
    Acute_Heart_Failure, Acute_Heart_Failure_Years,
    Cardiac_Arrest, Cardiac_Arrest_Years,
    Heart_Transplant, Heart_Transplant_Years,
    Ventricular_Tachycardia, Ventricular_Tachycardia_Years,
    Ventricular_Fib_Flutter_Tachy, Ventricular_Fib_Flutter_Tachy_Years,
    Defibrillator, Defibrillator_Years,
    AFib_Flutter, AFib_Flutter_Years,
    Septal_Reduction_Surgery, Septal_Reduction_Surgery_Years
  )

baseline_modeling_data <- baseline_cohort %>%
  mutate(
    Sex = case_when(
      Sex == 0 ~ "Male",
      Sex == 1 ~ "Female",
      TRUE ~ NA_character_
    ),
    ventricular_arrhythmia_composite = pmax(
      dplyr::coalesce(post_ventricular_tachycardia, 0),
      dplyr::coalesce(post_ventricular_fib_flut_tachy, 0),
      na.rm = TRUE
    ),
    ventricular_arrhythmia_yrs = pmin(
      if_else(post_ventricular_tachycardia == 1, post_ventricular_tachycardia_yrs, Inf),
      if_else(post_ventricular_fib_flut_tachy == 1, post_ventricular_fib_flut_tachy_yrs, Inf),
      na.rm = TRUE
    ),
    primary_composite = pmax(
      dplyr::coalesce(death, 0),
      dplyr::coalesce(post_cardiac_arrest, 0),
      dplyr::coalesce(post_heart_transplant, 0),
      na.rm = TRUE
    ),
    primary_composite_yrs = pmin(
      if_else(death == 1, death_yrs, Inf),
      if_else(post_cardiac_arrest == 1, post_cardiac_arrest_yrs, Inf),
      if_else(post_heart_transplant == 1, post_heart_transplant_yrs, Inf),
      na.rm = TRUE
    ),
    ventricular_arrhythmia_yrs = ifelse(is.infinite(ventricular_arrhythmia_yrs), NA_real_, ventricular_arrhythmia_yrs),
    primary_composite_yrs = ifelse(is.infinite(primary_composite_yrs), NA_real_, primary_composite_yrs),
    follow_up_yrs = as.numeric(last_enc_date - cpx_test_date) / 365.25
  ) %>%
  filter(!is.na(death)) %>%
  arrange(MRN, cpx_test_date) %>%
  mutate(
    across(
      c(
        death, post_acute_heart_failure, post_cardiac_arrest, post_heart_transplant,
        post_ventricular_tachycardia, post_ventricular_fib_flut_tachy, post_defibrillator,
        post_afib_flut, post_septal_reduction_surgery, primary_composite,
        ventricular_arrhythmia_composite, hf_dx, phtn_dx
      ),
      ~ case_when(.x == 1 ~ "Yes", .x == 0 ~ "No", TRUE ~ NA_character_)
    )
  ) %>%
  rename(
    Patient_ID = ID,
    Baseline_CPX_Date = cpx_test_date,
    Baseline_Test_Sequence = CPX.sequential,
    Age_Years = age,
    Peak_RER = pk.RER,
    Peak_VO2_mL_kg_min = VO2_kg,
    Peak_VO2_FRIEND1_pct = VO2_FRIEND_PP,
    Peak_VO2_FRIEND2_pct = VO2_FRIEND2_PP,
    Peak_VO2_Wasserman2005_pct = VO2_WASSERMAN_PP,
    Peak_HR_pct_predicted = HRmax_PP,
    HR_Recovery_1min = HRR,
    VE_VCO2_Slope = VeVco2_slope,
    Bike_Watts = `Bike.Watts`,
    Last_Encounter_Date = last_enc_date,
    Death = death,
    Death_Years = death_yrs,
    Acute_Heart_Failure = post_acute_heart_failure,
    Acute_Heart_Failure_Years = post_acute_heart_failure_yrs,
    Cardiac_Arrest = post_cardiac_arrest,
    Cardiac_Arrest_Years = post_cardiac_arrest_yrs,
    Heart_Transplant = post_heart_transplant,
    Heart_Transplant_Years = post_heart_transplant_yrs,
    Ventricular_Tachycardia = post_ventricular_tachycardia,
    Ventricular_Tachycardia_Years = post_ventricular_tachycardia_yrs,
    Ventricular_Fib_Flutter_Tachy = post_ventricular_fib_flut_tachy,
    Ventricular_Fib_Flutter_Tachy_Years = post_ventricular_fib_flut_tachy_yrs,
    Defibrillator = post_defibrillator,
    Defibrillator_Years = post_defibrillator_yrs,
    AFib_Flutter = post_afib_flut,
    AFib_Flutter_Years = post_afib_flut_yrs,
    Septal_Reduction_Surgery = post_septal_reduction_surgery,
    Septal_Reduction_Surgery_Years = post_septal_reduction_surgery_yrs,
    Heart_Failure_PreTest = hf_dx,
    Pulmonary_Hypertension_PreTest = phtn_dx,
    Ventricular_Arrhythmia_Composite = ventricular_arrhythmia_composite,
    Ventricular_Arrhythmia_Composite_Years = ventricular_arrhythmia_yrs,
    Primary_Composite = primary_composite,
    Primary_Composite_Years = primary_composite_yrs,
    FollowUp_Years = follow_up_yrs
  ) %>%
  select(
    Patient_ID, MRN, Baseline_CPX_Date, Baseline_Test_Sequence, Disease_group,
    Age_Years, Sex, Exercise_mode, Cycling_form, Protocol,
    BMI, BSA, LBM, LBMI, Peak_RER, Peak_VO2_mL_kg_min,
    Peak_VO2_FRIEND1_pct, Peak_VO2_FRIEND2_pct, Peak_VO2_Wasserman2005_pct,
    Peak_HR_pct_predicted, HR_Recovery_1min, VE_VCO2_Slope, OUES, Bike_Watts,
    Arrhythmia, Notes, Heart_Failure_PreTest, Pulmonary_Hypertension_PreTest,
    Last_Encounter_Date, FollowUp_Years,
    Primary_Composite, Primary_Composite_Years,
    Ventricular_Arrhythmia_Composite, Ventricular_Arrhythmia_Composite_Years,
    Death, Death_Years,
    Acute_Heart_Failure, Acute_Heart_Failure_Years,
    Cardiac_Arrest, Cardiac_Arrest_Years,
    Heart_Transplant, Heart_Transplant_Years,
    Ventricular_Tachycardia, Ventricular_Tachycardia_Years,
    Ventricular_Fib_Flutter_Tachy, Ventricular_Fib_Flutter_Tachy_Years,
    Defibrillator, Defibrillator_Years,
    AFib_Flutter, AFib_Flutter_Years,
    Septal_Reduction_Surgery, Septal_Reduction_Surgery_Years
  )

export_summary_overall <- tibble::tibble(
  Metric = c(
    "Export date",
    "Linked cycling CPX tests",
    "Unique patients",
    "Baseline modeling cohort",
    "Median follow-up, years",
    "Primary composite events",
    "Deaths"
  ),
  Value = c(
    format(Sys.Date()),
    nrow(linked_cycle_tests),
    dplyr::n_distinct(linked_cycle_tests$MRN),
    nrow(baseline_modeling_data),
    format(round(median(linked_cycle_tests$FollowUp_Years, na.rm = TRUE), 2), nsmall = 2),
    sum(linked_cycle_tests$Primary_Composite == "Yes", na.rm = TRUE),
    sum(linked_cycle_tests$Death == "Yes", na.rm = TRUE)
  )
)

export_summary_by_group <- linked_cycle_tests %>%
  count(Disease_group, name = "Tests") %>%
  left_join(
    linked_cycle_tests %>%
      group_by(Disease_group) %>%
      summarise(
        Patients = dplyr::n_distinct(MRN),
        Primary_Composite_Events = sum(Primary_Composite == "Yes", na.rm = TRUE),
        Deaths = sum(Death == "Yes", na.rm = TRUE),
        .groups = "drop"
      ),
    by = "Disease_group"
  )

export_dictionary <- tibble::tibble(
  Sheet = c(rep("Baseline Modeling Cohort", length(names(baseline_modeling_data))), rep("Linked Tests", length(names(linked_cycle_tests)))),
  Column = c(names(baseline_modeling_data), names(linked_cycle_tests)),
  Description = c(
    "Stable patient identifier generated from MRN within the cycle cohort",
    "Medical record number",
    "First linked ergocycle CPX date for the patient",
    "Sequential CPX number for the baseline test",
    "Phenotype assignment used in the cycle report",
    "Age at baseline CPX, years",
    "Biologic sex from source sheet",
    "Exercise mode field from source sheet",
    "Derived cycling form",
    "Protocol label from source sheet",
    "Body mass index, kg/m^2",
    "Body surface area, m^2",
    "Lean body mass, kg",
    "Lean body mass index, kg/m^2",
    "Peak respiratory exchange ratio",
    "Peak VO2, mL/kg/min",
    "Peak VO2 percent predicted, FRIEND 1",
    "Peak VO2 percent predicted, FRIEND 2",
    "Peak VO2 percent predicted, Wasserman 2005",
    "Peak heart rate percent predicted",
    "Heart rate recovery at 1 minute",
    "VE/VCO2 slope",
    "Oxygen uptake efficiency slope",
    "Peak bike workload, watts",
    "Arrhythmia field from source sheet",
    "Notes from source sheet",
    "Pre-test heart failure flag from comorbidity sheet",
    "Pre-test pulmonary hypertension flag from comorbidity sheet",
    "Last encounter date from outcomes sheet",
    "Observed follow-up from baseline CPX to last encounter, years",
    "Composite of death, cardiac arrest, or transplant",
    "Years from baseline CPX to primary composite event",
    "Composite of ventricular tachycardia or ventricular fibrillation/flutter/tachycardia",
    "Years from baseline CPX to ventricular arrhythmia composite event",
    "Death after baseline CPX",
    "Years from baseline CPX to death",
    "Acute heart failure hospitalization after baseline CPX",
    "Years from baseline CPX to acute heart failure",
    "Cardiac arrest after baseline CPX",
    "Years from baseline CPX to cardiac arrest",
    "Heart transplant after baseline CPX",
    "Years from baseline CPX to heart transplant",
    "Ventricular tachycardia after baseline CPX",
    "Years from baseline CPX to ventricular tachycardia",
    "Ventricular fibrillation/flutter/tachycardia after baseline CPX",
    "Years from baseline CPX to ventricular fibrillation/flutter/tachycardia",
    "Defibrillator after baseline CPX",
    "Years from baseline CPX to defibrillator",
    "Atrial fibrillation/flutter after baseline CPX",
    "Years from baseline CPX to atrial fibrillation/flutter",
    "Septal reduction surgery after baseline CPX",
    "Years from baseline CPX to septal reduction surgery",
    "Stable patient identifier generated from MRN within the cycle cohort",
    "Medical record number",
    "CPX test date",
    "Sequential CPX number from source sheet",
    "Phenotype assignment used in the cycle report",
    "Age at CPX, years",
    "Biologic sex from source sheet",
    "Exercise mode field from source sheet",
    "Derived cycling form",
    "Protocol label from source sheet",
    "Body mass index, kg/m^2",
    "Body surface area, m^2",
    "Lean body mass, kg",
    "Lean body mass index, kg/m^2",
    "Peak respiratory exchange ratio",
    "Peak VO2, mL/kg/min",
    "Peak VO2 percent predicted, FRIEND 1",
    "Peak VO2 percent predicted, FRIEND 2",
    "Peak VO2 percent predicted, Wasserman 2005",
    "Peak heart rate percent predicted",
    "Heart rate recovery at 1 minute",
    "VE/VCO2 slope",
    "Oxygen uptake efficiency slope",
    "Peak bike workload, watts",
    "Arrhythmia field from source sheet",
    "Notes from source sheet",
    "Pre-test heart failure flag from comorbidity sheet",
    "Pre-test pulmonary hypertension flag from comorbidity sheet",
    "Last encounter date from outcomes sheet",
    "Observed follow-up from CPX to last encounter, years",
    "Composite of death, cardiac arrest, or transplant",
    "Years from CPX to primary composite event",
    "Composite of ventricular tachycardia or ventricular fibrillation/flutter/tachycardia",
    "Years from CPX to ventricular arrhythmia composite event",
    "Death after CPX",
    "Years from CPX to death",
    "Acute heart failure hospitalization after CPX",
    "Years from CPX to acute heart failure",
    "Cardiac arrest after CPX",
    "Years from CPX to cardiac arrest",
    "Heart transplant after CPX",
    "Years from CPX to heart transplant",
    "Ventricular tachycardia after CPX",
    "Years from CPX to ventricular tachycardia",
    "Ventricular fibrillation/flutter/tachycardia after CPX",
    "Years from CPX to ventricular fibrillation/flutter/tachycardia",
    "Defibrillator after CPX",
    "Years from CPX to defibrillator",
    "Atrial fibrillation/flutter after CPX",
    "Years from CPX to atrial fibrillation/flutter",
    "Septal reduction surgery after CPX",
    "Years from CPX to septal reduction surgery"
  )
)

wb_export <- openxlsx::createWorkbook(creator = "OpenAI Codex")
openxlsx::addWorksheet(wb_export, "Summary", gridLines = FALSE)
openxlsx::addWorksheet(wb_export, "Baseline Modeling Cohort", gridLines = FALSE)
openxlsx::addWorksheet(wb_export, "Linked Tests", gridLines = FALSE)
openxlsx::addWorksheet(wb_export, "Data Dictionary", gridLines = FALSE)

title_style <- openxlsx::createStyle(
  fontSize = 14, textDecoration = "bold", fgFill = "#DCE6F1",
  halign = "left", border = "Bottom", borderColour = "#4F81BD"
)
section_style <- openxlsx::createStyle(
  fontSize = 11, textDecoration = "bold", fgFill = "#EAF2F8",
  border = "Bottom", borderColour = "#A9CCE3"
)
header_style <- openxlsx::createStyle(
  textDecoration = "bold", fgFill = "#1F4E78", fontColour = "#FFFFFF",
  halign = "center", valign = "center", border = "TopBottomLeftRight",
  borderColour = "#FFFFFF"
)
body_style <- openxlsx::createStyle(valign = "top")
date_style <- openxlsx::createStyle(numFmt = "yyyy-mm-dd")
decimal_style <- openxlsx::createStyle(numFmt = "0.00")
integer_style <- openxlsx::createStyle(numFmt = "0")
wrap_style <- openxlsx::createStyle(valign = "top", wrapText = TRUE)

openxlsx::writeData(wb_export, "Summary", x = "Ergocycle CPX linked outcomes export", startCol = 1, startRow = 1)
openxlsx::addStyle(wb_export, "Summary", title_style, rows = 1, cols = 1, gridExpand = TRUE)
openxlsx::writeData(wb_export, "Summary", x = "Overall", startCol = 1, startRow = 3)
openxlsx::addStyle(wb_export, "Summary", section_style, rows = 3, cols = 1, gridExpand = TRUE)
openxlsx::writeData(wb_export, "Summary", x = export_summary_overall, startCol = 1, startRow = 4, headerStyle = header_style)
openxlsx::writeData(wb_export, "Summary", x = "By phenotype", startCol = 1, startRow = 12)
openxlsx::addStyle(wb_export, "Summary", section_style, rows = 12, cols = 1, gridExpand = TRUE)
openxlsx::writeData(wb_export, "Summary", x = export_summary_by_group, startCol = 1, startRow = 13, headerStyle = header_style)
openxlsx::setColWidths(wb_export, "Summary", cols = 1:5, widths = c(28, 16, 14, 24, 12))
openxlsx::freezePane(wb_export, "Summary", firstActiveRow = 4)

openxlsx::writeDataTable(
  wb_export, "Baseline Modeling Cohort", x = baseline_modeling_data,
  tableStyle = "TableStyleMedium9", withFilter = TRUE
)
openxlsx::addStyle(
  wb_export, "Baseline Modeling Cohort", body_style,
  rows = 2:(nrow(baseline_modeling_data) + 1),
  cols = 1:ncol(baseline_modeling_data), gridExpand = TRUE, stack = TRUE
)

baseline_date_cols <- which(names(baseline_modeling_data) %in% c("Baseline_CPX_Date", "Last_Encounter_Date"))
baseline_decimal_cols <- which(names(baseline_modeling_data) %in% c(
  "BMI", "BSA", "LBM", "LBMI", "Peak_RER", "Peak_VO2_mL_kg_min",
  "Peak_VO2_FRIEND1_pct", "Peak_VO2_FRIEND2_pct", "Peak_VO2_Wasserman2005_pct",
  "Peak_HR_pct_predicted", "HR_Recovery_1min", "VE_VCO2_Slope", "OUES",
  "FollowUp_Years", "Primary_Composite_Years", "Ventricular_Arrhythmia_Composite_Years",
  "Death_Years", "Acute_Heart_Failure_Years", "Cardiac_Arrest_Years",
  "Heart_Transplant_Years", "Ventricular_Tachycardia_Years",
  "Ventricular_Fib_Flutter_Tachy_Years", "Defibrillator_Years",
  "AFib_Flutter_Years", "Septal_Reduction_Surgery_Years"
))
baseline_integer_cols <- which(names(baseline_modeling_data) %in% c("Patient_ID", "Baseline_Test_Sequence", "Age_Years", "Bike_Watts"))
baseline_wrap_cols <- which(names(baseline_modeling_data) %in% c("Protocol", "Arrhythmia", "Notes"))

if (length(baseline_date_cols) > 0) {
  openxlsx::addStyle(
    wb_export, "Baseline Modeling Cohort", date_style,
    rows = 2:(nrow(baseline_modeling_data) + 1), cols = baseline_date_cols,
    gridExpand = TRUE, stack = TRUE
  )
}
if (length(baseline_decimal_cols) > 0) {
  openxlsx::addStyle(
    wb_export, "Baseline Modeling Cohort", decimal_style,
    rows = 2:(nrow(baseline_modeling_data) + 1), cols = baseline_decimal_cols,
    gridExpand = TRUE, stack = TRUE
  )
}
if (length(baseline_integer_cols) > 0) {
  openxlsx::addStyle(
    wb_export, "Baseline Modeling Cohort", integer_style,
    rows = 2:(nrow(baseline_modeling_data) + 1), cols = baseline_integer_cols,
    gridExpand = TRUE, stack = TRUE
  )
}
if (length(baseline_wrap_cols) > 0) {
  openxlsx::addStyle(
    wb_export, "Baseline Modeling Cohort", wrap_style,
    rows = 2:(nrow(baseline_modeling_data) + 1), cols = baseline_wrap_cols,
    gridExpand = TRUE, stack = TRUE
  )
}

openxlsx::setColWidths(wb_export, "Baseline Modeling Cohort", cols = 1:ncol(baseline_modeling_data), widths = "auto")
if (length(baseline_wrap_cols) > 0) {
  openxlsx::setColWidths(wb_export, "Baseline Modeling Cohort", cols = baseline_wrap_cols, widths = 30)
}
openxlsx::freezePane(wb_export, "Baseline Modeling Cohort", firstActiveRow = 2, firstActiveCol = 6)

openxlsx::writeDataTable(
  wb_export, "Linked Tests", x = linked_cycle_tests,
  tableStyle = "TableStyleMedium2", withFilter = TRUE
)
openxlsx::addStyle(
  wb_export, "Linked Tests", body_style,
  rows = 2:(nrow(linked_cycle_tests) + 1),
  cols = 1:ncol(linked_cycle_tests), gridExpand = TRUE, stack = TRUE
)

date_cols <- which(names(linked_cycle_tests) %in% c("Test_Date", "Last_Encounter_Date"))
decimal_cols <- which(names(linked_cycle_tests) %in% c(
  "BMI", "BSA", "LBM", "LBMI", "Peak_RER", "Peak_VO2_mL_kg_min",
  "Peak_VO2_FRIEND1_pct", "Peak_VO2_FRIEND2_pct", "Peak_VO2_Wasserman2005_pct",
  "Peak_HR_pct_predicted", "HR_Recovery_1min", "VE_VCO2_Slope", "OUES",
  "FollowUp_Years", "Primary_Composite_Years", "Ventricular_Arrhythmia_Composite_Years",
  "Death_Years", "Acute_Heart_Failure_Years", "Cardiac_Arrest_Years",
  "Heart_Transplant_Years", "Ventricular_Tachycardia_Years",
  "Ventricular_Fib_Flutter_Tachy_Years", "Defibrillator_Years",
  "AFib_Flutter_Years", "Septal_Reduction_Surgery_Years"
))
integer_cols <- which(names(linked_cycle_tests) %in% c("Patient_ID", "Test_Sequence", "age", "Bike_Watts"))
wrap_cols <- which(names(linked_cycle_tests) %in% c("Protocol", "Arrhythmia", "Notes"))

if (length(date_cols) > 0) {
  openxlsx::addStyle(
    wb_export, "Linked Tests", date_style,
    rows = 2:(nrow(linked_cycle_tests) + 1), cols = date_cols,
    gridExpand = TRUE, stack = TRUE
  )
}
if (length(decimal_cols) > 0) {
  openxlsx::addStyle(
    wb_export, "Linked Tests", decimal_style,
    rows = 2:(nrow(linked_cycle_tests) + 1), cols = decimal_cols,
    gridExpand = TRUE, stack = TRUE
  )
}
if (length(integer_cols) > 0) {
  openxlsx::addStyle(
    wb_export, "Linked Tests", integer_style,
    rows = 2:(nrow(linked_cycle_tests) + 1), cols = integer_cols,
    gridExpand = TRUE, stack = TRUE
  )
}
if (length(wrap_cols) > 0) {
  openxlsx::addStyle(
    wb_export, "Linked Tests", wrap_style,
    rows = 2:(nrow(linked_cycle_tests) + 1), cols = wrap_cols,
    gridExpand = TRUE, stack = TRUE
  )
}

openxlsx::setColWidths(wb_export, "Linked Tests", cols = 1:ncol(linked_cycle_tests), widths = "auto")
if (length(wrap_cols) > 0) {
  openxlsx::setColWidths(wb_export, "Linked Tests", cols = wrap_cols, widths = 30)
}
openxlsx::freezePane(wb_export, "Linked Tests", firstActiveRow = 2, firstActiveCol = 6)

openxlsx::writeDataTable(
  wb_export, "Data Dictionary", x = export_dictionary,
  tableStyle = "TableStyleLight9", withFilter = TRUE
)
openxlsx::setColWidths(wb_export, "Data Dictionary", cols = 1:3, widths = c(24, 34, 88))
openxlsx::addStyle(
  wb_export, "Data Dictionary", wrap_style,
  rows = 2:(nrow(export_dictionary) + 1), cols = 1:3,
  gridExpand = TRUE, stack = TRUE
)
openxlsx::freezePane(wb_export, "Data Dictionary", firstActiveRow = 2)

openxlsx::saveWorkbook(wb_export, output_file("Cycle_Linked_Tests.xlsx"), overwrite = TRUE)

tibble::tibble(
  workbook = output_file("Cycle_Linked_Tests.xlsx"),
  linked_tests = nrow(linked_cycle_tests),
  unique_patients = dplyr::n_distinct(linked_cycle_tests$MRN),
  baseline_modeling_patients = nrow(baseline_modeling_data)
)
```

```{r cohort_flow_fig, fig.cap="Figure 1. Cohort derivation for ergocycle-based CPX analyses."}
ggplot(cohort_flow, aes(y = stage, x = n_tests, fill = group_label)) +
  geom_col(width = 0.68, color = "black", linewidth = 0.25) +
  geom_text(
    aes(label = comma(n_tests)),
    position = position_stack(vjust = 0.5),
    size = 3.2,
    color = "white",
    fontface = "bold"
  ) +
  geom_text(
    data = cohort_flow %>% distinct(stage, total_tests),
    aes(
      y = stage,
      x = total_tests + max(cohort_flow$total_tests) * 0.03,
      label = paste0(comma(total_tests), " tests")
    ),
    inherit.aes = FALSE,
    hjust = 0,
    size = 3.5
  ) +
  scale_fill_manual(
    values = group_display_colors,
    breaks = c("CON", "CHF", "PH"),
    labels = c("CON", "CHF", "PH"),
    name = "Phenotype"
  ) +
  scale_x_continuous(
    labels = comma,
    expand = expansion(mult = c(0, 0)),
    limits = c(0, max(cohort_flow$total_tests) * 1.35)
  ) +
  labs(x = "Observations", y = NULL) +
  theme_publication() +
  theme(legend.position = "right")
```

## Baseline phenotype

Baseline descriptive analyses used the first available ergocycle CPX for each patient. Patients were classified as `No HF/PHTN`, `HF`, or `PHTN` according to pre-test comorbidity flags; when both heart failure and pulmonary hypertension were present, the patient was assigned to the `PHTN` stratum to preserve a distinct pulmonary hypertension comparison group.

```{r baseline}
baseline_data <- baseline_cohort %>%
  mutate(
    Sex = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
    Cycling_form = factor(Cycling_form),
    across(
      c(age, BMI, BSA, LBMI, pk.RER, VO2_kg, VO2_FRIEND2_PP, VO2_WASSERMAN_PP, HRmax_PP, HRR, VeVco2_slope, OUES),
      ~ as.numeric(as.character(.))
    )
  )

table1 <- baseline_data %>%
  select(
    Disease_group, age, Sex, Cycling_form, BMI, BSA, LBMI, pk.RER,
    VO2_kg, VO2_FRIEND2_PP, VO2_WASSERMAN_PP, HRmax_PP, HRR, VeVco2_slope, OUES
  ) %>%
  tbl_summary(
    by = Disease_group,
    label = list(
      age ~ "Age, years",
      Sex ~ "Sex",
      Cycling_form ~ "Cycling form",
      BMI ~ "Body mass index, kg/m^2",
      BSA ~ "Body surface area, m^2",
      LBMI ~ "Lean body mass index, kg/m^2",
      pk.RER ~ "Peak RER",
      VO2_kg ~ "Peak VO2, mL/kg/min",
      VO2_FRIEND2_PP ~ "Peak VO2, % predicted (FRIEND 2.0)",
      VO2_WASSERMAN_PP ~ "Peak VO2, % predicted (Wasserman 2005)",
      HRmax_PP ~ "Peak heart rate, % predicted",
      HRR ~ "Heart rate recovery at 1 minute",
      VeVco2_slope ~ "VE/VCO2 slope",
      OUES ~ "OUES"
    ),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1,
    missing = "no"
  ) %>%
  add_overall(last = TRUE) %>%
  bold_labels() %>%
  modify_caption("**Table 1. Baseline characteristics stratified by heart failure and pulmonary hypertension phenotype**")

table1
table1 %>%
  as_flex_table() %>%
  save_as_docx(path = output_file("Table1_Cycle_Baseline.docx"))
```

```{r baseline_figure, fig.cap="Figure 2. Baseline ergocycle CPX phenotype stratified by heart failure and pulmonary hypertension status."}
group_labels <- baseline_data %>%
  count(Disease_group) %>%
  mutate(label = paste0(Disease_group, "\n(n=", n, ")")) %>%
  select(Disease_group, label)

baseline_plot_data <- baseline_data %>%
  left_join(group_labels, by = "Disease_group") %>%
  mutate(label = factor(label, levels = group_labels$label))

make_group_boxplot <- function(data, yvar, ylab, title_text) {
  ggplot(data, aes(x = label, y = .data[[yvar]], fill = Disease_group, color = Disease_group)) +
    geom_violin(alpha = 0.20, width = 0.9, trim = FALSE) +
    geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.85, color = "black") +
    geom_jitter(width = 0.10, alpha = 0.30, size = 1.1, stroke = 0) +
    scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors) +
    labs(title = title_text, x = NULL, y = ylab) +
    theme_publication() +
    theme(legend.position = "none")
}

p_age <- make_group_boxplot(baseline_plot_data, "age", "Years", "Age")
p_vo2 <- make_group_boxplot(
  baseline_plot_data, "VO2_FRIEND2_PP",
  expression("% predicted"), expression("Peak " * VO[2] ~ "(FRIEND 2.0)")
)
p_vevco2 <- make_group_boxplot(
  baseline_plot_data, "VeVco2_slope",
  expression(VE/VCO[2] ~ "slope"), expression(VE/VCO[2] ~ "slope")
)
p_hrr <- make_group_boxplot(
  baseline_plot_data, "HRR",
  "Beats/min", "Heart rate recovery"
)

(p_age | p_vo2) / (p_vevco2 | p_hrr)

ggsave(
  filename = output_file("Figure2_Cycle_Baseline.pdf"),
  plot = (p_age | p_vo2) / (p_vevco2 | p_hrr),
  width = 11,
  height = 8,
  units = "in",
  device = "pdf",
  useDingbats = FALSE
)
```

## Post-CPX clinical outcomes

Clinical outcomes were summarized from the outcomes sheet using the first ergocycle CPX as baseline. The primary endpoint was a composite of death, cardiac arrest, or heart transplant. Secondary endpoints were ventricular arrhythmia composite and the individual outcomes of death, acute heart failure, cardiac arrest, and heart transplant.

```{r outcomes}
baseline_outcomes <- baseline_data %>%
  transmute(
    ID,
    Disease_group,
    cpx_test_date,
    age,
    Sex,
    VO2_FRIEND2_PP,
    VeVco2_slope,
    HRR,
    last_enc_date,
    death,
    death_yrs,
    acute_heart_failure = post_acute_heart_failure,
    acute_heart_failure_yrs = post_acute_heart_failure_yrs,
    cardiac_arrest = post_cardiac_arrest,
    cardiac_arrest_yrs = post_cardiac_arrest_yrs,
    heart_transplant = post_heart_transplant,
    heart_transplant_yrs = post_heart_transplant_yrs,
    ventricular_arrhythmia_composite = pmax(
      dplyr::coalesce(post_ventricular_tachycardia, 0),
      dplyr::coalesce(post_ventricular_fib_flut_tachy, 0),
      na.rm = TRUE
    ),
    ventricular_arrhythmia_yrs = pmin(
      if_else(post_ventricular_tachycardia == 1, post_ventricular_tachycardia_yrs, Inf),
      if_else(post_ventricular_fib_flut_tachy == 1, post_ventricular_fib_flut_tachy_yrs, Inf),
      na.rm = TRUE
    ),
    primary_composite = pmax(
      dplyr::coalesce(death, 0),
      dplyr::coalesce(post_cardiac_arrest, 0),
      dplyr::coalesce(post_heart_transplant, 0),
      na.rm = TRUE
    ),
    primary_composite_yrs = pmin(
      if_else(death == 1, death_yrs, Inf),
      if_else(post_cardiac_arrest == 1, post_cardiac_arrest_yrs, Inf),
      if_else(post_heart_transplant == 1, post_heart_transplant_yrs, Inf),
      na.rm = TRUE
    )
  ) %>%
  mutate(
    ventricular_arrhythmia_yrs = ifelse(is.infinite(ventricular_arrhythmia_yrs), NA_real_, ventricular_arrhythmia_yrs),
    primary_composite_yrs = ifelse(is.infinite(primary_composite_yrs), NA_real_, primary_composite_yrs),
    follow_up_yrs = as.numeric(last_enc_date - cpx_test_date) / 365.25
  )

outcome_definitions <- tibble::tribble(
  ~outcome, ~event_col, ~time_col,
  "Primary composite", "primary_composite", "primary_composite_yrs",
  "Ventricular arrhythmia composite", "ventricular_arrhythmia_composite", "ventricular_arrhythmia_yrs",
  "Death", "death", "death_yrs",
  "Acute heart failure", "acute_heart_failure", "acute_heart_failure_yrs",
  "Cardiac arrest", "cardiac_arrest", "cardiac_arrest_yrs",
  "Heart transplant", "heart_transplant", "heart_transplant_yrs"
)

summarize_outcome <- function(data, group_value, outcome_label, event_col, time_col) {
  sub <- if (identical(group_value, "Overall")) data else dplyr::filter(data, Disease_group == group_value)
  events <- sub[[event_col]]
  times <- sub[[time_col]]
  tibble::tibble(
    group = group_value,
    outcome = outcome_label,
    n = sum(events == 1, na.rm = TRUE),
    pct = 100 * mean(events == 1, na.rm = TRUE),
    median_years = ifelse(sum(events == 1, na.rm = TRUE) > 0, median(times[events == 1], na.rm = TRUE), NA_real_)
  )
}

outcome_summary <- dplyr::bind_rows(
  lapply(seq_len(nrow(outcome_definitions)), function(i) {
    def <- outcome_definitions[i, ]
    dplyr::bind_rows(
      summarize_outcome(baseline_outcomes, "Overall", def$outcome, def$event_col, def$time_col),
      lapply(levels(baseline_outcomes$Disease_group), function(g) {
        summarize_outcome(baseline_outcomes, g, def$outcome, def$event_col, def$time_col)
      })
    )
  })
) %>%
  mutate(outcome = factor(outcome, levels = outcome_definitions$outcome))

outcome_table <- baseline_outcomes %>%
  transmute(
    Disease_group,
    primary_composite = factor(primary_composite, levels = c(1, 0), labels = c("Yes", "No")),
    ventricular_arrhythmia_composite = factor(ventricular_arrhythmia_composite, levels = c(1, 0), labels = c("Yes", "No")),
    death = factor(death, levels = c(1, 0), labels = c("Yes", "No")),
    acute_heart_failure = factor(acute_heart_failure, levels = c(1, 0), labels = c("Yes", "No")),
    cardiac_arrest = factor(cardiac_arrest, levels = c(1, 0), labels = c("Yes", "No")),
    heart_transplant = factor(heart_transplant, levels = c(1, 0), labels = c("Yes", "No"))
  ) %>%
  tbl_summary(
    by = Disease_group,
    label = list(
      primary_composite ~ "Primary composite",
      ventricular_arrhythmia_composite ~ "Ventricular arrhythmia composite",
      death ~ "Death",
      acute_heart_failure ~ "Acute heart failure",
      cardiac_arrest ~ "Cardiac arrest",
      heart_transplant ~ "Heart transplant"
    ),
    statistic = all_categorical() ~ "{n} ({p}%)",
    missing = "no"
  ) %>%
  add_p(
    test = all_categorical() ~ "fisher.test",
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) %>%
  add_overall(last = TRUE) %>%
  bold_labels() %>%
  modify_caption("**Table 2. Post-CPX clinical outcomes stratified by heart failure and pulmonary hypertension phenotype**")

outcome_table
outcome_table %>%
  as_flex_table() %>%
  save_as_docx(path = output_file("Table2_Cycle_Outcomes.docx"))
```

```{r outcomes_figure, fig.cap="Figure 3. Baseline phenotype and subsequent clinical outcomes."}
outcome_plot_data <- outcome_summary %>%
  filter(group != "Overall") %>%
  mutate(group = factor(group, levels = c("No HF/PHTN", "HF", "PHTN")))

p_outcomes <- ggplot(outcome_plot_data, aes(x = outcome, y = pct, fill = group)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.66) +
  geom_text(
    aes(label = sprintf("%d", n)),
    position = position_dodge(width = 0.72),
    vjust = -0.2,
    size = 3
  ) +
  scale_fill_manual(values = group_colors) +
  coord_flip() +
  labs(x = NULL, y = "Patients with event (%)") +
  theme_publication()

p_time <- ggplot(
  outcome_plot_data %>% filter(!is.na(median_years)),
  aes(x = outcome, y = median_years, fill = group)
) +
  geom_col(position = position_dodge(width = 0.72), width = 0.66) +
  scale_fill_manual(values = group_colors) +
  coord_flip() +
  labs(x = NULL, y = "Median time to event (years)") +
  theme_publication()

p_outcomes | p_time

ggsave(
  output_file("Figure3_Cycle_Outcomes.pdf"),
  p_outcomes | p_time,
  width = 12,
  height = 7,
  units = "in",
  device = "pdf"
)
```

## Time-to-event analysis

Kaplan-Meier analyses were performed from the baseline ergocycle CPX to the first occurrence of the endpoint or censoring at last recorded follow-up. Curves are stratified by baseline phenotype, and accompanying statistics include log-rank tests and Cox model hazard ratios using `No HF/PHTN` as the reference group.

```{r km_analysis}
build_surv_dataset <- function(data, event_col, time_col) {
  data %>%
    transmute(
      ID,
      Disease_group,
      event = as.numeric(.data[[event_col]]),
      event_time = as.numeric(.data[[time_col]]),
      censor_time = as.numeric(follow_up_yrs)
    ) %>%
    mutate(
      time = if_else(event == 1 & is.finite(event_time), event_time, censor_time),
      status = if_else(event == 1 & is.finite(event_time), 1, 0)
    ) %>%
    filter(is.finite(time), time >= 0, !is.na(Disease_group))
}

tidy_survfit <- function(fit) {
  surv_summary <- summary(fit)
  surv_df <- tibble::tibble(
    time = surv_summary$time,
    surv = surv_summary$surv,
    std.err = surv_summary$std.err,
    lower = surv_summary$lower,
    upper = surv_summary$upper,
    strata = gsub("^Disease_group=", "", surv_summary$strata)
  )

  start_df <- tibble::tibble(
    time = 0,
    surv = 1,
    std.err = 0,
    lower = 1,
    upper = 1,
    strata = levels(baseline_outcomes$Disease_group)
  )

  bind_rows(start_df, surv_df) %>%
    mutate(strata = factor(strata, levels = levels(baseline_outcomes$Disease_group)))
}

km_stats <- function(data, endpoint_label) {
  surv_obj <- survival::Surv(data$time, data$status)
  fit <- survival::survfit(surv_obj ~ Disease_group, data = data)
  logrank <- survival::survdiff(surv_obj ~ Disease_group, data = data)
  logrank_p <- 1 - pchisq(logrank$chisq, length(logrank$n) - 1)
  cox_fit <- survival::coxph(surv_obj ~ Disease_group, data = data)
  cox_tidy <- summary(cox_fit)$coefficients
  cox_ci <- summary(cox_fit)$conf.int

  tibble::tibble(
    Endpoint = endpoint_label,
    N = nrow(data),
    Events = sum(data$status == 1, na.rm = TRUE),
    `Log-rank p` = format.pval(logrank_p, digits = 2, eps = 0.001),
    `HR HF vs CON` = sprintf(
      "%.2f (%.2f-%.2f)",
      cox_ci["Disease_groupHF", "exp(coef)"],
      cox_ci["Disease_groupHF", "lower .95"],
      cox_ci["Disease_groupHF", "upper .95"]
    ),
    `HR PH vs CON` = sprintf(
      "%.2f (%.2f-%.2f)",
      cox_ci["Disease_groupPHTN", "exp(coef)"],
      cox_ci["Disease_groupPHTN", "lower .95"],
      cox_ci["Disease_groupPHTN", "upper .95"]
    )
  )
}

primary_surv <- build_surv_dataset(baseline_outcomes, "primary_composite", "primary_composite_yrs")
vent_surv <- build_surv_dataset(baseline_outcomes, "ventricular_arrhythmia_composite", "ventricular_arrhythmia_yrs")

km_statistics <- bind_rows(
  km_stats(primary_surv, "Primary composite"),
  km_stats(vent_surv, "Ventricular arrhythmia composite")
)

km_statistics_ft <- flextable::flextable(km_statistics)
flextable::save_as_docx(km_statistics_ft, path = output_file("Table3_Cycle_KM_Statistics.docx"))

km_statistics
```

```{r km_figure, fig.cap="Figure 4. Kaplan-Meier curves for primary and secondary time-to-event outcomes."}
make_km_plot <- function(data, endpoint_label) {
  fit <- survival::survfit(survival::Surv(time, status) ~ Disease_group, data = data)
  plot_data <- tidy_survfit(fit)
  logrank <- survival::survdiff(survival::Surv(time, status) ~ Disease_group, data = data)
  logrank_p <- 1 - pchisq(logrank$chisq, length(logrank$n) - 1)

  ggplot(plot_data, aes(x = time, y = surv, color = strata, fill = strata)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.12, linewidth = 0) +
    geom_step(linewidth = 1.1) +
    scale_color_manual(values = group_colors, labels = group_display) +
    scale_fill_manual(values = group_colors, labels = group_display) +
    scale_x_continuous(limits = c(0, km_max_years), breaks = seq(0, km_max_years, by = 2)) +
    scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
    labs(
      title = endpoint_label,
      subtitle = paste0("Log-rank p ", format.pval(logrank_p, digits = 2, eps = 0.001)),
      x = "Years from baseline CPX",
      y = "Event-free survival",
      color = "Phenotype",
      fill = "Phenotype"
    ) +
    theme_publication() +
    theme(legend.position = "right")
}

p_km_primary <- make_km_plot(primary_surv, "Primary composite")
p_km_vent <- make_km_plot(vent_surv, "Ventricular arrhythmia composite")

km_figure <- p_km_primary | p_km_vent
km_figure

ggsave(
  output_file("Figure4_Cycle_KM.pdf"),
  km_figure,
  width = 12.5,
  height = 5.5,
  units = "in",
  device = "pdf"
)
```
