---
title: "Modeling Left Ventricular Diastolic Dysfunction and Cardiopulmonary Fitness in HCM"
author: "Mark E. Pepin, MD, PhD, MS, FESC"
date: "2026-03-25"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-tools: true
    code-fold: true
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 6.9,
  fig.height = 4.5,
  dpi = 150,
  dev = "png",
  dev.args = list(type = "cairo")
)

# ── Package loading ──────────────────────────────────────────────────────────
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(patchwork)
  library(splines)
  library(lme4)
  library(mgcv)
  library(broom)
  library(survival)
  library(scales)
  library(stringr)
  library(openxlsx)
  library(gtsummary)
  library(flextable)
  library(ggcorrplot)
  library(viridis)
  library(ggrepel)
  library(gamlss)
  library(reticulate)
})

reticulate::use_python("/Users/pepin/.pyenv/versions/3.12.12/bin/python3", required = TRUE)

# ── JACC color palette (colorblind-safe) ─────────────────────────────────────
jacc_cols <- c(
  navy    = "#2C3E50",
  red     = "#E74C3C",
  blue    = "#3498DB",
  green   = "#27AE60",
  orange  = "#F39C12",
  gray    = "#7F8C8D",
  purple  = "#8E44AD"
)

# ── JACC publication theme ───────────────────────────────────────────────────
theme_jacc <- function(base_size = 9) {
  theme_classic(base_size = base_size, base_family = "Arial") +
    theme(
      plot.title       = element_text(face = "bold", size = base_size + 1, hjust = 0.5),
      plot.subtitle    = element_text(size = base_size, hjust = 0.5, color = "grey40"),
      axis.title       = element_text(face = "bold"),
      axis.text        = element_text(color = "black"),
      axis.line        = element_line(color = "black", linewidth = 0.4),
      axis.ticks       = element_line(color = "black", linewidth = 0.3),
      panel.grid       = element_blank(),
      strip.background = element_rect(fill = "grey94", color = NA),
      strip.text       = element_text(face = "bold", size = base_size),
      legend.position  = "top",
      legend.title     = element_blank(),
      legend.text      = element_text(size = base_size - 1),
      plot.margin      = margin(4, 8, 4, 4)
    )
}

# ── Export helper ────────────────────────────────────────────────────────────
save_jacc <- function(plot, path, w = 6.9, h = 4.5) {
  ggsave(filename = path, plot = plot, device = cairo_pdf,
         width = w, height = h, units = "in", dpi = 600)
}

# ── Age-residualization helper (single definition) ───────────────────────────
residualize_on_age <- function(df, idx) {
  df <- df %>% mutate(.row_id = row_number())
  dsub <- df %>% filter(!is.na(.data[[idx]]), !is.na(age))
  if (nrow(dsub) < 10) {
    df[[paste0(idx, "_resid")]] <- NA_real_
    return(df %>% select(-.row_id))
  }
  fit <- lm(as.formula(paste0(idx, " ~ ns(age, df = 3)")), data = dsub)
  dsub$.resid <- resid(fit)
  df <- df %>% left_join(dsub %>% select(.row_id, .resid), by = ".row_id")
  df[[paste0(idx, "_resid")]] <- df$.resid
  df %>% select(-.row_id, -.resid)
}

# ── Output directory ─────────────────────────────────────────────────────────
dir.create("../2_Output", recursive = TRUE, showWarnings = FALSE)
```

# Data Import and Cohort Assembly

## Import and clean CPX data

The initial phase imports raw Cardiopulmonary Exercise Testing (CPET) data, assigns a stable patient identifier (ID) per MRN, retains patients with serial testing ($\geq 2$ sessions), restricts to HCM with adequate effort (peak RER $\geq 1.0$), and derives biometric indices (BSA, BMI, LBMI).

```{r cpx_cleaning}
CPX <- read.xlsx("../1_Input/CPX_input.xlsx", sheet = "CPX", detectDates = TRUE)

CPX_with_id <- CPX %>%
  group_by(MRN) %>%
  mutate(ID = cur_group_id()) %>%
  ungroup()

CPX_serial <- CPX_with_id %>%
  group_by(ID) %>%
  filter(n() > 1) %>%
  ungroup()

CPX_hcm_effort <- CPX_serial %>%
  filter(ID != "1380") %>%
  filter(MCl3 == "HCM", pk.RER >= 1.0)

# Cohort flow tracking
filter_flow <- tibble(
  stage      = c("Initial CPX import", "Assigned stable patient ID",
                 "Retained serial CPX only", "HCM diagnosis + RER >= 1.0"),
  n_tests    = c(nrow(CPX), nrow(CPX_with_id), nrow(CPX_serial), nrow(CPX_hcm_effort)),
  n_patients = c(n_distinct(CPX$MRN), n_distinct(CPX_with_id$ID),
                 n_distinct(CPX_serial$ID), n_distinct(CPX_hcm_effort$ID))
)

# Biometric calculations
CPX_clean <- CPX_hcm_effort %>%
  mutate(
    BSA  = sqrt((`height.(cm)` * `Weight.(kg)`) / 3600),
    LBMI = LBM.NHANES.no.race / (`height.(cm)` / 100)^2,
    BMI  = `Weight.(kg)` / ((`height.(cm)` / 100)^2)
  ) %>%
  group_by(ID) %>%
  arrange(cpx_test_date, .by_group = TRUE) %>%
  mutate(days_from_baseline = as.numeric(cpx_test_date - first(cpx_test_date))) %>%
  ungroup()

CPX_clean <- CPX_clean %>%
  select(
    ID, MRN, cpx_test_date, age, Sex = `Sex:.M(0)/F(1)`, BMI, BSA,
    LBM = LBM.NHANES.no.race, LBMI, HCM_Phenotype = MCl4,
    CPX.sequential, days_from_baseline, pk.RER,
    VO2_FRIEND_PP   = `FRIEND1.%.predicted.VO2`,
    VO2_WASSERMAN_PP = `Wasserman.%predicted.VO2.(2005)`,
    VO2_FRIEND2_PP  = `FRIEND2.%predicted.VO2`,
    HRmax_PP = `%predicted.HR.FRIEND`,
    HRR = `HR.recovery.(1min)`,
    VeVco2_slope = `ve/vco2.slope`
  ) %>%
  filter(
    (is.na(VO2_FRIEND2_PP) | (VO2_FRIEND2_PP >= 0 & VO2_FRIEND2_PP <= 200)) &
    (is.na(VO2_FRIEND_PP)  | (VO2_FRIEND_PP >= 0 & VO2_FRIEND_PP <= 200))
  )

cat(sprintf("CPX tests (HCM, adequate effort): %d tests, %d patients\n",
            nrow(CPX_clean), n_distinct(CPX_clean$ID)))
```

## Comorbidity exclusions

Patients with documented CAD, COPD, or interstitial lung disease are excluded so that exercise limitations are attributable to HCM-specific pathology.

```{r comorbidities}
Comorbidities_table <- read.xlsx("../1_Input/CPX_input.xlsx",
                                  sheet = "co-morbidities", detectDates = TRUE)

Comorbidities_to_filter <- Comorbidities_table %>%
  select(MRN, cad_pre_test, copd_pre_test, interstitial_lung_dz_pre_test) %>%
  filter(
    (cad_pre_test != 1 | is.na(cad_pre_test)) &
    (copd_pre_test != 1 | is.na(copd_pre_test)) &
    (interstitial_lung_dz_pre_test != 1 | is.na(interstitial_lung_dz_pre_test))
  )

CPX_comorbidity_filtered <- CPX_clean %>%
  filter(MRN %in% Comorbidities_to_filter$MRN)

filter_flow <- bind_rows(filter_flow, tibble(
  stage      = "Excluded CAD/COPD/ILD",
  n_tests    = nrow(CPX_comorbidity_filtered),
  n_patients = n_distinct(CPX_comorbidity_filtered$ID)
))
```

## Merge echocardiographic data

Each CPET is aligned with the closest prior resting echocardiogram within a configurable window (default 365 days).

```{r echo_merge}
echo_raw <- read.xlsx("../1_Input/mark_extract_20260316.xlsx", sheet = 1, detectDates = TRUE)

echo_vars_core <- c(
  "e_e_ave", "e_e_lat", "e_e_med", "mv_a_dur", "mv_a_point", "mv_dec_time",
  "mv_e_a", "tr_max_vel", "la_vol_index", "med_peak_e_vel", "lat_peak_e_vel",
  "ef_modsp4", "la_vol", "pulm_dias_vel", "pulm_sys_vel"
)

echo_all <- echo_raw %>%
  mutate(MRN = as.character(mrn), echo_date = as.Date(echo_date)) %>%
  select(MRN, echo_date, all_of(echo_vars_core)) %>%
  mutate(across(all_of(echo_vars_core), ~ as.numeric(as.character(.))))

# Closest prior echo within window
align_window_days <- 365

cpx_aligned <- CPX_comorbidity_filtered %>%
  mutate(MRN = as.character(MRN), cpx_test_date = as.Date(cpx_test_date))

cpx_echo <- cpx_aligned %>%
  left_join(echo_all, by = "MRN", relationship = "many-to-many") %>%
  mutate(delta_days = as.numeric(cpx_test_date - echo_date)) %>%
  filter(!is.na(delta_days), delta_days >= 0, delta_days <= align_window_days) %>%
  group_by(ID, CPX.sequential, cpx_test_date) %>%
  slice_min(order_by = delta_days, n = 1, with_ties = FALSE) %>%
  ungroup()

# Prepare derived variables
cpx_echo <- cpx_echo %>%
  mutate(
    E_vel = coalesce(med_peak_e_vel, lat_peak_e_vel),
    e_prime_ave = ifelse(!is.na(med_peak_e_vel) & !is.na(lat_peak_e_vel),
                         (med_peak_e_vel + lat_peak_e_vel) / 2,
                         coalesce(med_peak_e_vel, lat_peak_e_vel)),
    Sex   = factor(Sex, levels = c(0, 1), labels = c("Male", "Female"))
  )

# Update filter flow
filter_flow <- bind_rows(filter_flow, tibble(
  stage      = "Echo aligned (within 365d)",
  n_tests    = nrow(cpx_echo),
  n_patients = n_distinct(cpx_echo$ID)
))

# Baseline and longitudinal cohorts
num_vars <- c("VO2_FRIEND2_PP", "VO2_FRIEND_PP", "VO2_WASSERMAN_PP", "VeVco2_slope",
              "pk.RER", "age", "BMI", "LBMI", "e_e_ave", "la_vol_index", "tr_max_vel",
              "med_peak_e_vel", "lat_peak_e_vel", "ef_modsp4",
              "mv_e_a", "mv_dec_time", "E_vel", "e_prime_ave")

baseline_df <- cpx_echo %>%
  group_by(ID) %>% arrange(days_from_baseline) %>% slice(1) %>% ungroup() %>%
  mutate(across(any_of(num_vars), ~ as.numeric(as.character(.))))

long_df <- cpx_echo %>%
  filter(!is.na(VO2_FRIEND2_PP), !is.na(days_from_baseline)) %>%
  mutate(across(any_of(num_vars), ~ as.numeric(as.character(.))),
         time_yrs = days_from_baseline / 365.25)

cat(sprintf("Baseline cohort: %d patients\nLongitudinal observations: %d\n",
            nrow(baseline_df), nrow(long_df)))
```

## Import clinical outcomes

```{r outcomes_import}
outcomes_raw <- read.xlsx("../1_Input/CPX_input.xlsx", sheet = "Outcomes_2025", detectDates = TRUE)

outcomes <- outcomes_raw %>%
  mutate(MRN = as.character(MRN)) %>%
  select(MRN,
         death, death_date_clean,
         post_acute_heart_failure, post_acute_heart_failure_date, post_acute_heart_failure_yrs,
         post_chronic_heart_failure, post_chronic_heart_failure_date, post_chronic_heart_failure_yrs,
         post_heart_transplant, post_heart_transplant_date, post_heart_transplant_yrs,
         post_afib_flut, post_afib_flut_date, post_afib_flut_yrs,
         last_enc_date) %>%
  mutate(across(c(death, post_acute_heart_failure, post_chronic_heart_failure,
                  post_heart_transplant, post_afib_flut), ~ as.numeric(as.character(.)))) %>%
  mutate(across(c(post_acute_heart_failure_yrs, post_chronic_heart_failure_yrs,
                  post_heart_transplant_yrs, post_afib_flut_yrs), ~ as.numeric(as.character(.))))

# Composite HF endpoint: acute HF, chronic HF, transplant, or death
outcomes <- outcomes %>%
  mutate(
    hf_composite = pmax(
      coalesce(post_acute_heart_failure, 0),
      coalesce(post_chronic_heart_failure, 0),
      coalesce(post_heart_transplant, 0),
      coalesce(death, 0),
      na.rm = TRUE
    ),
    hf_composite_yrs = pmin(
      coalesce(post_acute_heart_failure_yrs, Inf),
      coalesce(post_chronic_heart_failure_yrs, Inf),
      coalesce(post_heart_transplant_yrs, Inf),
      na.rm = TRUE
    ),
    hf_composite_yrs = ifelse(is.infinite(hf_composite_yrs), NA_real_, hf_composite_yrs)
  )

# Merge to baseline
baseline_df <- baseline_df %>%
  left_join(outcomes %>% select(MRN, hf_composite, hf_composite_yrs,
                                 post_afib_flut, post_afib_flut_yrs,
                                 death, last_enc_date),
            by = "MRN")

cat(sprintf("Patients with outcome data: %d\nComposite HF events: %d\n",
            sum(!is.na(baseline_df$hf_composite)),
            sum(baseline_df$hf_composite == 1, na.rm = TRUE)))
```

## Cohort flow diagram

```{r flow_diagram, fig.width=6.9, fig.height=3.8, results='asis'}
filter_flow <- filter_flow %>%
  mutate(
    stage = factor(stage, levels = stage),
    removed = lag(n_tests, default = first(n_tests)) - n_tests
  )

p_flow <- filter_flow %>%
  mutate(stage = factor(stage, levels = rev(levels(stage)))) %>%
  ggplot(aes(y = stage, x = n_tests)) +
  geom_col(fill = jacc_cols["navy"], width = 0.65, color = "black", linewidth = 0.2) +
  geom_text(aes(label = paste0(comma(n_tests), " tests | ", comma(n_patients), " pts")),
            hjust = 0, nudge_x = 15, size = 2.8, color = "black") +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.25))) +
  labs(title = "Cohort Assembly", x = "CPX Tests", y = NULL) +
  theme_jacc()

p_flow
save_jacc(p_flow, "../2_Output/Cohort_Filtering_Workflow.pdf", w = 6.9, h = 3.8)
```

# ASE 2025 Filling Pressure Classification (HCM-Specific Algorithm)

The 2025 ASE guidelines (Nagueh et al., *JASE* 2025) provide a disease-specific algorithm for patients with known myocardial disease, including HCM. In contrast to the general-population algorithm that proceeds through a two-step hierarchy (Step 1: impaired relaxation screening → Step 2: filling pressure grading into Grades I–III), the HCM-specific pathway **presumes impaired myocardial relaxation** given the underlying myopathy, hypertrophy, and fiber disarray, and proceeds directly to filling pressure estimation. The resulting classification is **binary**: Normal vs. Elevated LV filling pressures.

Three primary criteria are evaluated for concordance: average E/e$'$ >14, left atrial volume index (LAVI) >34 mL/m$^2$, and peak tricuspid regurgitation velocity (TR V$_{max}$) >2.8 m/s. When all three are concordantly normal or abnormal, classification is straightforward. When discordant, tiebreaker criteria are applied: left atrial reservoir strain (LARS $\leq$ 18%), pulmonary venous Ar-A duration difference ($\geq$ 30 ms), and E/A ratio ($\leq$ 0.8 with E velocity $\leq$ 50 cm/s [normal] or $\geq$ 2 [elevated]). LARS and Ar-A duration were not available in this dataset; the E/A ratio serves as the primary tiebreaker.

```{r ase_grading, results='asis'}
# ── ASE 2025 HCM-Specific Filling Pressure Algorithm ──────────────────────────
# In HCM, impaired relaxation is PRESUMED → skip Step 1 → classify directly by
# filling pressure markers into: Normal vs Elevated
classify_filling_pressure <- function(e_e_ave, la_vol_index, tr_max_vel,
                                       mv_e_a, E_vel) {

  # ── Three primary concordance criteria ─────────────────────────────────────
  # Note: tr_max_vel is stored in cm/s; ASE cutoff is >2.8 m/s = >280 cm/s
  abn_ee   <- !is.na(e_e_ave)      & e_e_ave > 14
  abn_lavi <- !is.na(la_vol_index)  & la_vol_index > 34
  abn_trv  <- !is.na(tr_max_vel)    & tr_max_vel > 280

  avail_ee   <- !is.na(e_e_ave)
  avail_lavi <- !is.na(la_vol_index)
  avail_trv  <- !is.na(tr_max_vel)

  n_primary_available <- avail_ee + avail_lavi + avail_trv
  n_primary_abnormal  <- abn_ee + abn_lavi + abn_trv

  # ── Concordance logic ──────────────────────────────────────────────────────
  # All 3 concordantly normal → Normal FP

  # All 3 concordantly abnormal → Elevated FP
  # Discordant → apply tiebreakers
  concordant_normal   <- (n_primary_available >= 2) & (n_primary_abnormal == 0)
  concordant_elevated <- (n_primary_available >= 2) & (n_primary_abnormal == n_primary_available)

  # ── Tiebreaker: E/A ratio (LARS and Ar-A unavailable) ─────────────────────
  # E/A ≤ 0.8 AND E ≤ 50 cm/s → favors normal LAP (normal FP)
  # E/A ≥ 2 → favors elevated FP
  ea_favors_normal   <- !is.na(mv_e_a) & !is.na(E_vel) &
                         mv_e_a <= 0.8 & E_vel <= 50
  ea_favors_elevated <- !is.na(mv_e_a) & mv_e_a >= 2

  # ── Classification ─────────────────────────────────────────────────────────
  fp_class <- case_when(
    n_primary_available < 2                               ~ "Indeterminate",
    concordant_normal                                     ~ "Normal",
    concordant_elevated                                   ~ "Elevated",
    # Discordant: majority rule with tiebreaker
    n_primary_abnormal >= 2                               ~ "Elevated",
    n_primary_abnormal == 0                               ~ "Normal",
    # 1 of 3 abnormal (or 1 of 2 abnormal) → use tiebreaker
    n_primary_abnormal == 1 & ea_favors_elevated          ~ "Elevated",
    n_primary_abnormal == 1 & ea_favors_normal            ~ "Normal",
    # Tiebreaker unavailable or equivocal
    TRUE                                                  ~ "Indeterminate"
  )

  factor(fp_class, levels = c("Normal", "Elevated", "Indeterminate"),
         ordered = FALSE)
}

baseline_df <- baseline_df %>%
  mutate(fp_class = classify_filling_pressure(
    e_e_ave, la_vol_index, tr_max_vel, mv_e_a, E_vel
  ))

long_df <- long_df %>%
  left_join(
    baseline_df %>% select(ID, fp_class),
    by = "ID"
  )

# ── Concordance summary ──────────────────────────────────────────────────────
baseline_df <- baseline_df %>%
  mutate(
    abn_ee   = !is.na(e_e_ave)      & e_e_ave > 14,
    abn_lavi = !is.na(la_vol_index)  & la_vol_index > 34,
    abn_trv  = !is.na(tr_max_vel)    & tr_max_vel > 280,
    n_primary_available = (!is.na(e_e_ave)) + (!is.na(la_vol_index)) + (!is.na(tr_max_vel)),
    n_primary_abnormal  = abn_ee + abn_lavi + abn_trv,
    concordant = (n_primary_abnormal == 0) | (n_primary_abnormal == n_primary_available),
    discordant = n_primary_available >= 2 & !concordant
  )

cat("**ASE 2025 HCM-Specific Filling Pressure Classification:**\n\n")
cat("  In HCM, impaired myocardial relaxation is PRESUMED.\n")
cat("  Classification proceeds directly to filling pressure estimation.\n\n")

cat("**Primary Criteria Concordance:**\n\n")
cat(sprintf("  Concordant cases:  %d (%0.1f%%)\n",
            sum(baseline_df$concordant & baseline_df$n_primary_available >= 2, na.rm = TRUE),
            100 * mean(baseline_df$concordant[baseline_df$n_primary_available >= 2], na.rm = TRUE)))
cat(sprintf("  Discordant cases:  %d (%0.1f%%)\n\n",
            sum(baseline_df$discordant, na.rm = TRUE),
            100 * mean(baseline_df$discordant[baseline_df$n_primary_available >= 2], na.rm = TRUE)))

cat("**Filling Pressure Classification (Baseline):**\n\n")
print(table(baseline_df$fp_class, useNA = "ifany"))

cat(sprintf("\n\n  Normal:        %d (%0.1f%%)\n",
            sum(baseline_df$fp_class == "Normal", na.rm = TRUE),
            100 * mean(baseline_df$fp_class == "Normal", na.rm = TRUE)))
cat(sprintf("  Elevated:      %d (%0.1f%%)\n",
            sum(baseline_df$fp_class == "Elevated", na.rm = TRUE),
            100 * mean(baseline_df$fp_class == "Elevated", na.rm = TRUE)))
cat(sprintf("  Indeterminate: %d (%0.1f%%)\n",
            sum(baseline_df$fp_class == "Indeterminate", na.rm = TRUE),
            100 * mean(baseline_df$fp_class == "Indeterminate", na.rm = TRUE)))

# Individual criterion prevalence
cat("\n**Individual Criterion Prevalence:**\n\n")
cat(sprintf("  E/e' >14:     %d / %d (%0.1f%%)\n",
            sum(baseline_df$abn_ee, na.rm = TRUE),
            sum(!is.na(baseline_df$e_e_ave)),
            100 * mean(baseline_df$abn_ee[!is.na(baseline_df$e_e_ave)], na.rm = TRUE)))
cat(sprintf("  LAVI >34:     %d / %d (%0.1f%%)\n",
            sum(baseline_df$abn_lavi, na.rm = TRUE),
            sum(!is.na(baseline_df$la_vol_index)),
            100 * mean(baseline_df$abn_lavi[!is.na(baseline_df$la_vol_index)], na.rm = TRUE)))
cat(sprintf("  TR Vmax >2.8: %d / %d (%0.1f%%)\n",
            sum(baseline_df$abn_trv, na.rm = TRUE),
            sum(!is.na(baseline_df$tr_max_vel)),
            100 * mean(baseline_df$abn_trv[!is.na(baseline_df$tr_max_vel)], na.rm = TRUE)))
```

```{r ase_figure, fig.width=6.9, fig.height=5.0, results='asis'}
# ── Panel A: Filling pressure classification distribution ─────────────────────
df_fp <- baseline_df %>% filter(fp_class != "Indeterminate")

p_bar <- df_fp %>%
  count(fp_class) %>%
  ggplot(aes(x = fp_class, y = n, fill = fp_class)) +
  geom_col(width = 0.6, color = "black", linewidth = 0.2) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Normal" = jacc_cols["blue"],
                                "Elevated" = jacc_cols["red"])) +
  labs(title = "LV Filling Pressure Classification",
       x = "ASE 2025 (HCM-Specific)", y = "Patients") +
  theme_jacc() + theme(legend.position = "none")

# ── Panel B: Individual criterion concordance heatmap ────────────────────────
df_criteria <- baseline_df %>%
  filter(fp_class != "Indeterminate") %>%
  select(fp_class, `E/e' >14` = abn_ee, `LAVI >34` = abn_lavi, `TR Vmax >2.8` = abn_trv) %>%
  pivot_longer(-fp_class, names_to = "Criterion", values_to = "Abnormal") %>%
  group_by(fp_class, Criterion) %>%
  summarise(pct = 100 * mean(Abnormal, na.rm = TRUE), .groups = "drop")

p_criteria <- ggplot(df_criteria, aes(x = Criterion, y = fp_class, fill = pct)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.0f%%", pct)), size = 3.5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = jacc_cols["red"],
                      limits = c(0, 100), name = "% Abnormal") +
  labs(title = "Criterion Prevalence by FP Group",
       x = NULL, y = NULL) +
  theme_jacc() + theme(legend.position = "right",
                        axis.text.x = element_text(angle = 25, hjust = 1))

# ── Panel C: Peak VO2 by filling pressure ─────────────────────────────────────
p_box <- df_fp %>%
  filter(!is.na(VO2_FRIEND2_PP)) %>%
  ggplot(aes(x = fp_class, y = VO2_FRIEND2_PP, fill = fp_class)) +
  geom_boxplot(width = 0.5, outlier.size = 0.8, linewidth = 0.3) +
  geom_jitter(width = 0.15, alpha = 0.15, size = 0.8) +
  scale_fill_manual(values = c("Normal" = jacc_cols["blue"],
                                "Elevated" = jacc_cols["red"])) +
  annotate("text", x = 1.5, y = Inf, vjust = 1.5, size = 3, fontface = "italic",
           label = paste0("p = ", format.pval(
             wilcox.test(VO2_FRIEND2_PP ~ fp_class, data = df_fp)$p.value, digits = 2))) +
  labs(title = bquote(bold("Peak" ~ VO[2] ~ "by Filling Pressure")),
       x = "LV Filling Pressure",
       y = bquote("Peak" ~ VO[2] ~ "(% Predicted, FRIEND 2.0)")) +
  theme_jacc() + theme(legend.position = "none")

# ── Panel D: VE/VCO2 by filling pressure ─────────────────────────────────────
p_vevco2 <- df_fp %>%
  filter(!is.na(VeVco2_slope)) %>%
  ggplot(aes(x = fp_class, y = VeVco2_slope, fill = fp_class)) +
  geom_boxplot(width = 0.5, outlier.size = 0.8, linewidth = 0.3) +
  geom_jitter(width = 0.15, alpha = 0.15, size = 0.8) +
  scale_fill_manual(values = c("Normal" = jacc_cols["blue"],
                                "Elevated" = jacc_cols["red"])) +
  annotate("text", x = 1.5, y = Inf, vjust = 1.5, size = 3, fontface = "italic",
           label = paste0("p = ", format.pval(
             wilcox.test(VeVco2_slope ~ fp_class, data = df_fp)$p.value, digits = 2))) +
  labs(title = bquote(bold(VE/VCO[2] ~ "Slope by Filling Pressure")),
       x = "LV Filling Pressure",
       y = bquote(VE/VCO[2] ~ "Slope")) +
  theme_jacc() + theme(legend.position = "none")

fig_ase <- (p_bar | p_criteria) / (p_box | p_vevco2) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "bold", size = 11))

fig_ase
save_jacc(fig_ase, "../2_Output/Figure_ASE2025_FillingPressure.pdf", w = 6.9, h = 5.0)
```

```{r ase_stratified_table, results='asis'}
tbl_ase <- baseline_df %>%
  filter(fp_class != "Indeterminate") %>%
  select(fp_class, age, Sex, BMI, LBMI, HCM_Phenotype,
         VO2_FRIEND2_PP, VO2_WASSERMAN_PP, VeVco2_slope, HRR,
         e_prime_ave, med_peak_e_vel, lat_peak_e_vel,
         e_e_ave, la_vol_index, tr_max_vel, mv_e_a, mv_dec_time, ef_modsp4) %>%
  tbl_summary(
    by = fp_class,
    label = list(
      age              ~ "Age, years",
      Sex              ~ "Sex",
      BMI              ~ "BMI, kg/m\u00B2",
      LBMI             ~ "LBMI, kg/m\u00B2",
      HCM_Phenotype    ~ "HCM Phenotype",
      VO2_FRIEND2_PP   ~ "Peak VO\u2082 (FRIEND 2.0 %pred)",
      VO2_WASSERMAN_PP ~ "Peak VO\u2082 (Wasserman %pred)",
      VeVco2_slope     ~ "VE/VCO\u2082 Slope",
      HRR              ~ "HR Recovery (1 min)",
      e_prime_ave      ~ "Average e', cm/s",
      med_peak_e_vel   ~ "Septal e', cm/s",
      lat_peak_e_vel   ~ "Lateral e', cm/s",
      e_e_ave          ~ "E/e' (average)",
      la_vol_index     ~ "LAVI, mL/m\u00B2",
      tr_max_vel       ~ "TR Vmax, cm/s",
      mv_e_a           ~ "E/A Ratio",
      mv_dec_time      ~ "MV Deceleration Time, ms",
      ef_modsp4        ~ "LVEF, %"
    ),
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 1,
    missing = "no"
  ) %>%
  add_p() %>%
  bold_labels() %>%
  modify_caption("**Table: Patient Characteristics by LV Filling Pressure (ASE 2025 HCM Algorithm)**")

tbl_ase

tbl_ase %>% as_flex_table() %>%
  save_as_docx(path = "../2_Output/Table_ASE2025_FillingPressure.docx")
```

# GAMLSS / SEAMLSS Continuous Modeling

We employ a Semiparametric Additive Model for Location, Scale, and Shape (GAMLSS) to model the continuous relationship between diastolic indices and Peak VO$_2$. The SEAMLSS framework derives Z-scores via $u(x) = \ln(y / \mu(x))$, where $\mu(x)$ is the fitted conditional mean, providing a stable relative metric that avoids the global distortion of Box-Cox (LMS) transformations.

```{r gamlss_modeling}
# Fit GAMLSS for each key diastolic index against Peak VO2
diastolic_key <- c("e_e_ave", "la_vol_index", "tr_max_vel", "e_prime_ave", "mv_e_a")
diastolic_labels <- c("e_e_ave"      = "E/e' (average)",
                       "la_vol_index" = "LAVI (mL/m\u00B2)",
                       "tr_max_vel"   = "TR Vmax (cm/s)",
                       "e_prime_ave"  = "Average e' (cm/s)",
                       "mv_e_a"       = "E/A Ratio")

gamlss_results <- list()

for (idx in diastolic_key) {
  df_fit <- baseline_df %>%
    filter(!is.na(.data[[idx]]), !is.na(VO2_FRIEND2_PP), !is.na(age), !is.na(LBMI)) %>%
    mutate(Sex_num = as.numeric(Sex == "Female"))

  if (nrow(df_fit) < 30) next

  # GAMLSS requires no NAs in the data frame — subset to used columns only
  gamlss_cols <- c("VO2_FRIEND2_PP", idx, "age", "Sex_num", "LBMI")
  df_gamlss <- df_fit[, gamlss_cols, drop = FALSE] %>% na.omit()
  if (nrow(df_gamlss) < 30) next

  tryCatch({
    frm_mu <- as.formula(paste0("VO2_FRIEND2_PP ~ cs(", idx, ", df=3) + cs(age, df=3) + Sex_num + LBMI"))
    m <- gamlss(
      frm_mu,
      sigma.formula = ~ cs(age, df=3),
      data = df_gamlss,
      family = NO(),
      control = gamlss.control(n.cyc = 100, trace = FALSE)
    )

    # SEAMLSS Z-score: u(x) = ln(y / mu(x))
    mu_hat <- fitted(m, what = "mu")
    sigma_hat <- fitted(m, what = "sigma")
    df_gamlss$seamlss_z <- log(df_gamlss$VO2_FRIEND2_PP / mu_hat)
    df_gamlss$gamlss_z  <- (df_gamlss$VO2_FRIEND2_PP - mu_hat) / sigma_hat
    df_gamlss$mu_hat    <- mu_hat

    # Merge back to df_fit (which has ID and other columns)
    df_out <- df_fit
    df_out$seamlss_z <- NA_real_; df_out$gamlss_z <- NA_real_; df_out$mu_hat <- NA_real_
    match_idx <- match(rownames(df_gamlss), rownames(df_fit))
    df_out$seamlss_z[match_idx] <- df_gamlss$seamlss_z
    df_out$gamlss_z[match_idx]  <- df_gamlss$gamlss_z
    df_out$mu_hat[match_idx]    <- df_gamlss$mu_hat

    gamlss_results[[idx]] <- list(model = m, data = df_out)
  }, error = function(e) {
    message(sprintf("GAMLSS for %s failed: %s", idx, e$message))
  })
}
```

```{r gamlss_figures, fig.width=6.9, fig.height=6.5}
if (length(gamlss_results) > 0) {
  # Panel plots: observed vs fitted centile curves
  centile_plots <- lapply(names(gamlss_results), function(idx) {
    res <- gamlss_results[[idx]]
    df  <- res$data

    ggplot(df, aes(x = .data[[idx]], y = VO2_FRIEND2_PP)) +
      geom_point(alpha = 0.3, size = 1, color = jacc_cols["gray"]) +
      geom_smooth(aes(y = mu_hat), color = jacc_cols["navy"], linewidth = 1,
                  method = "loess", se = FALSE) +
      geom_smooth(method = "loess", color = jacc_cols["red"], linewidth = 0.7,
                  linetype = "dashed", se = TRUE, alpha = 0.1) +
      labs(title = diastolic_labels[idx],
           x = diastolic_labels[idx],
           y = bquote("Peak" ~ VO[2] ~ "(% Pred)")) +
      theme_jacc()
  })

  # SEAMLSS Z-score distribution
  z_data <- bind_rows(lapply(names(gamlss_results), function(idx) {
    gamlss_results[[idx]]$data %>%
      select(seamlss_z) %>%
      mutate(index = diastolic_labels[idx])
  }))

  p_zdist <- ggplot(z_data, aes(x = seamlss_z, fill = index)) +
    geom_histogram(aes(y = after_stat(density)), bins = 30, alpha = 0.6, color = "white") +
    geom_density(aes(color = index), linewidth = 0.8) +
    facet_wrap(~ index, scales = "free") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(title = "SEAMLSS Z-Score Distribution",
         x = "Z-score: ln(observed / predicted)",
         y = "Density") +
    theme_jacc() + theme(legend.position = "none")

  # Combine
  n_plots <- length(centile_plots)
  top_row <- wrap_plots(centile_plots, nrow = 1)
  fig_gamlss <- top_row / p_zdist +
    plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(face = "bold", size = 11))

  fig_gamlss
  save_jacc(fig_gamlss, "../2_Output/Figure_GAMLSS_SEAMLSS.pdf", w = 6.9, h = 6.5)
}
```

```{r gamlss_table, results='asis'}
if (length(gamlss_results) > 0) {
  gamlss_summary <- bind_rows(lapply(names(gamlss_results), function(idx) {
    m <- gamlss_results[[idx]]$model
    # Capture summary output as a matrix
    s <- tryCatch({
      capture.output(sm <- summary(m))
      as.data.frame(sm) %>%
        tibble::rownames_to_column("term") %>%
        mutate(index = diastolic_labels[idx], AIC = AIC(m))
    }, error = function(e) {
      # Fallback: extract coefficients directly
      co <- coef(m)
      tibble(term = names(co), Estimate = co, index = diastolic_labels[idx], AIC = AIC(m))
    })
    s
  }))

  gamlss_summary %>%
    mutate(across(where(is.numeric), ~ round(., 4))) %>%
    flextable() %>%
    bold(part = "header") %>%
    theme_booktabs() %>%
    autofit() %>%
    set_caption("Table 2. GAMLSS Model Coefficients for Diastolic Indices vs Peak VO\u2082")

  write.csv(gamlss_summary,
            "../2_Output/Table2_GAMLSS_Coefficients.csv", row.names = FALSE)
}
```

# Cross-Sectional Nonlinear Analysis

Individual diastolic indices are modeled against Peak VO$_2$ using natural splines (df=3) adjusted for age, sex, and lean body mass index. Likelihood ratio tests compare each diastolic-augmented model to the clinical base model.

```{r cross_sectional_helpers, include=FALSE}
diastolic_indices <- c(
 "e_e_ave", "e_e_lat", "e_e_med", "mv_a_dur", "mv_a_point", "mv_dec_time",
 "mv_e_a", "tr_max_vel", "la_vol_index", "med_peak_e_vel", "lat_peak_e_vel",
 "e_prime_ave"
)

label_map <- c(
  e_e_ave        = "E/e' (avg)",    e_e_lat        = "E/e' (lat)",
  e_e_med        = "E/e' (med)",    mv_a_dur       = "MV A dur",
  mv_a_point     = "MV A point",    mv_dec_time    = "MV decel time",
  mv_e_a         = "MV E/A",        tr_max_vel     = "TR Vmax",
  la_vol_index   = "LAVI",          med_peak_e_vel = "Septal e'",
  lat_peak_e_vel = "Lateral e'",    e_prime_ave    = "Avg e'"
)

pred_grid <- function(var, df, model, lo, hi, n = 200) {
  g <- tibble(!!var := seq(lo, hi, length.out = n))
  ref_vars <- intersect(c("age", "BMI", "LBMI"), colnames(df))
  ref <- df %>% summarise(across(all_of(ref_vars), ~ median(., na.rm = TRUE)),
                           Sex = as.character(names(which.max(table(Sex)))))
  newd <- bind_cols(ref %>% slice(rep(1, n)), g)
  newd$Sex <- factor(newd$Sex, levels = levels(df$Sex))
  p <- predict(model, newdata = newd, se.fit = TRUE)
  tibble(x = newd[[var]], fit = p$fit, se = p$se.fit,
         lo = p$fit - 1.96 * p$se.fit, hi = p$fit + 1.96 * p$se.fit, var = var)
}
```

## Primary outcome: Peak VO$_2$ (FRIEND 2.0 % predicted)

```{r cross_sectional_friend2, fig.width=6.9, fig.height=6.5}
# Individual index models
baseline_models <- list()
baseline_lrt <- list()

for (idx in diastolic_indices) {
  df_idx <- baseline_df %>%
    filter(!is.na(.data[["VO2_FRIEND2_PP"]]), !is.na(age), !is.na(Sex),
           !is.na(LBMI), !is.na(.data[[idx]]))
  if (nrow(df_idx) < 20) next

  base <- lm(VO2_FRIEND2_PP ~ age + Sex + LBMI, data = df_idx)
  mod  <- lm(as.formula(paste0("VO2_FRIEND2_PP ~ age + Sex + LBMI + ns(", idx, ", df=3)")),
             data = df_idx)
  lrt  <- anova(base, mod)

  baseline_models[[idx]] <- list(df = df_idx, base = base, mod = mod, lrt = lrt)
  baseline_lrt[[idx]] <- tibble(
    index = idx, label = label_map[idx], n = nrow(df_idx),
    delta_AIC = AIC(mod) - AIC(base),
    F_stat = lrt$F[2], p_value = lrt$`Pr(>F)`[2],
    R2_base = summary(base)$r.squared, R2_full = summary(mod)$r.squared
  )
}

lrt_table <- bind_rows(baseline_lrt) %>% arrange(p_value)
write.csv(lrt_table, "../2_Output/Stats_CrossSectional_LRT_FRIEND2.csv", row.names = FALSE)

# Spline prediction curves
if (length(baseline_models) > 0) {
  pp_all <- bind_rows(lapply(names(baseline_models), function(idx) {
    m <- baseline_models[[idx]]
    lo <- quantile(m$df[[idx]], 0.02, na.rm = TRUE)
    hi <- quantile(m$df[[idx]], 0.98, na.rm = TRUE)
    pred_grid(idx, m$df, m$mod, lo, hi) %>% mutate(label = label_map[var])
  }))

  fig3 <- ggplot(pp_all, aes(x = x, y = fit)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, fill = jacc_cols["blue"]) +
    geom_line(linewidth = 0.9, color = jacc_cols["navy"]) +
    facet_wrap(~ label, scales = "free_x", nrow = 3) +
    labs(title = bquote(bold("Nonlinear Association: Diastolic Indices and Peak" ~ VO[2])),
         x = NULL,
         y = bquote("Predicted Peak" ~ VO[2] ~ "(FRIEND 2.0 % Pred)")) +
    theme_jacc()

  fig3
  save_jacc(fig3, "../2_Output/Figure3_Nonlinear_DiastolicIndices_FRIEND2.pdf", w = 6.9, h = 6.5)
}
```

```{r cross_sectional_lrt_table, results='asis'}
lrt_table %>%
  mutate(
    p_value = ifelse(p_value < 0.001, "<0.001", sprintf("%.3f", p_value)),
    across(c(delta_AIC, R2_base, R2_full, F_stat), ~ round(., 2))
  ) %>%
  flextable() %>%
  bold(part = "header") %>%
  theme_booktabs() %>%
  autofit() %>%
  set_caption("Likelihood Ratio Tests: Diastolic Indices vs Peak VO\u2082 (FRIEND 2.0)")
```

## Secondary outcome: VE/VCO$_2$ slope

```{r cross_sectional_vevco2, fig.width=6.9, fig.height=6.5}
vevco2_models <- list()
vevco2_lrt <- list()

for (idx in diastolic_indices) {
  df_idx <- baseline_df %>%
    filter(!is.na(VeVco2_slope), !is.na(age), !is.na(Sex),
           !is.na(LBMI), !is.na(.data[[idx]]))
  if (nrow(df_idx) < 20) next

  base <- lm(VeVco2_slope ~ age + Sex + LBMI, data = df_idx)
  mod  <- lm(as.formula(paste0("VeVco2_slope ~ age + Sex + LBMI + ns(", idx, ", df=3)")),
             data = df_idx)
  lrt  <- anova(base, mod)

  vevco2_models[[idx]] <- list(df = df_idx, base = base, mod = mod, lrt = lrt)
  vevco2_lrt[[idx]] <- tibble(
    index = idx, label = label_map[idx], n = nrow(df_idx),
    delta_AIC = AIC(mod) - AIC(base),
    F_stat = lrt$F[2], p_value = lrt$`Pr(>F)`[2]
  )
}

vevco2_lrt_table <- bind_rows(vevco2_lrt) %>% arrange(p_value)
write.csv(vevco2_lrt_table, "../2_Output/Stats_CrossSectional_LRT_VeVco2.csv", row.names = FALSE)

if (length(vevco2_models) > 0) {
  pp_vevco2 <- bind_rows(lapply(names(vevco2_models), function(idx) {
    m <- vevco2_models[[idx]]
    lo <- quantile(m$df[[idx]], 0.02, na.rm = TRUE)
    hi <- quantile(m$df[[idx]], 0.98, na.rm = TRUE)
    pred_grid(idx, m$df, m$mod, lo, hi) %>% mutate(label = label_map[var])
  }))

  fig_vevco2 <- ggplot(pp_vevco2, aes(x = x, y = fit)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, fill = jacc_cols["orange"]) +
    geom_line(linewidth = 0.9, color = jacc_cols["red"]) +
    facet_wrap(~ label, scales = "free_x", nrow = 3) +
    labs(title = bquote(bold("Nonlinear Association: Diastolic Indices and" ~ VE/VCO[2] ~ "Slope")),
         x = NULL,
         y = bquote(VE/VCO[2] ~ "Slope")) +
    theme_jacc()

  fig_vevco2
  save_jacc(fig_vevco2, "../2_Output/FigureS_Nonlinear_DiastolicIndices_VeVco2.pdf", w = 6.9, h = 6.5)
}
```

## Sensitivity: Echo alignment window

```{r sensitivity_window}
run_window <- function(w, idx, outcome_var = "VO2_FRIEND2_PP") {
  tmp <- cpx_aligned %>%
    left_join(echo_all, by = "MRN", relationship = "many-to-many") %>%
    mutate(delta_days = as.numeric(cpx_test_date - echo_date)) %>%
    filter(!is.na(delta_days), delta_days >= 0, delta_days <= w) %>%
    group_by(ID, CPX.sequential, cpx_test_date) %>%
    slice_min(order_by = delta_days, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      e_prime_ave = ifelse(!is.na(med_peak_e_vel) & !is.na(lat_peak_e_vel),
                           (med_peak_e_vel + lat_peak_e_vel) / 2,
                           coalesce(med_peak_e_vel, lat_peak_e_vel))
    ) %>%
    mutate(across(any_of(c(idx, "age", "LBMI", outcome_var)), ~ as.numeric(as.character(.))),
           Sex = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")))

  base <- tmp %>%
    group_by(ID) %>% arrange(days_from_baseline) %>% slice(1) %>% ungroup() %>%
    filter(!is.na(.data[[outcome_var]]), !is.na(.data[[idx]]), !is.na(age), !is.na(Sex), !is.na(LBMI))

  if (nrow(base) < 20) return(tibble(window_days = w, index = idx, n = nrow(base),
                                      delta_AIC = NA_real_, p_value = NA_real_))

  fit_base <- lm(as.formula(paste0(outcome_var, " ~ age + Sex + LBMI")), data = base)
  fit_idx  <- lm(as.formula(paste0(outcome_var, " ~ age + Sex + LBMI + ns(", idx, ", df=3)")), data = base)
  lrt <- anova(fit_base, fit_idx)

  tibble(window_days = w, index = idx, n = nrow(base),
         delta_AIC = AIC(fit_idx) - AIC(fit_base), p_value = lrt$`Pr(>F)`[2])
}

sens <- bind_rows(lapply(diastolic_indices, function(idx) {
  bind_rows(run_window(180, idx), run_window(365, idx), run_window(730, idx))
}))

write.csv(sens, "../2_Output/Sensitivity_EchoWindow_FRIEND2.csv", row.names = FALSE)
```

# OMAR: Adaptive Threshold Discovery

OMAR (Open Multivariate Adaptive Regression) is a modern, open-source implementation of the MARS (Multivariate Adaptive Regression Splines) algorithm. Unlike restricted cubic splines which require pre-specified knot locations, OMAR automatically discovers threshold breakpoints (hinge functions) in the data through a forward/backward stepwise procedure. This data-driven approach is particularly valuable for identifying clinically meaningful thresholds in the relationship between diastolic indices and cardiopulmonary fitness that may differ from guideline-derived cutoffs.

```{python omar_analysis, eval=TRUE}
import numpy as np
import pandas as pd
from omar import OMAR

# Import baseline data from R via reticulate
baseline_r = r.baseline_df

# Prepare analysis dataframe
omar_vars = ['e_e_ave', 'la_vol_index', 'tr_max_vel', 'e_prime_ave',
             'mv_e_a', 'mv_dec_time', 'age', 'LBMI']
outcome_vars = ['VO2_FRIEND2_PP', 'VeVco2_slope']
diastolic_vars = {'e_e_ave', 'la_vol_index', 'tr_max_vel', 'e_prime_ave', 'mv_e_a', 'mv_dec_time'}

df = baseline_r[omar_vars + outcome_vars + ['Sex']].copy()
df['Sex_num'] = pd.to_numeric(df['Sex'].map({'Male': 0, 'Female': 1}), errors='coerce')
df = df.drop(columns=['Sex'])

# Run OMAR for each outcome
omar_results = {}
for outcome in outcome_vars:
    predictors = [v for v in omar_vars if v != outcome] + ['Sex_num']
    df_complete = df[predictors + [outcome]].dropna()

    if len(df_complete) < 30:
        continue

    X_raw = df_complete[predictors].values.astype(np.float64)
    y = df_complete[outcome].values.astype(np.float64)

    # Standardize X to improve numerical stability (Cholesky in OMAR)
    X_mean = np.nanmean(X_raw, axis=0)
    X_std  = np.nanstd(X_raw, axis=0)
    X_std[X_std == 0] = 1.0
    X = (X_raw - X_mean) / X_std

    # OMAR API: max_nbases, max_ncandidates (<= max_nbases), penalty (GCV)
    try:
        model = OMAR(max_nbases=10, max_ncandidates=10, penalty=3.0)
        lof = model.find_bases(X, y)
    except Exception as e:
        print(f"  OMAR fitting failed for {outcome}: {e}")
        continue

    # Score: compute R²
    y_pred = model(X)
    ss_res = np.sum((y - y_pred)**2)
    ss_tot = np.sum((y - y.mean())**2)
    r2 = 1 - ss_res / ss_tot

    # Extract active basis functions and their thresholds
    # mask: 2D (max_nbases × max_nbases). Columns = basis functions. Rows = multiplicative factors.
    # cov: variable index for each factor. root: knot value. truncated: hinge flag.
    active_idx = np.where(np.any(model.mask, axis=0))[0]
    basis_info = []
    for j, bi in enumerate(active_idx):
        active_rows = np.where(model.mask[:, bi])[0]
        for row in active_rows:
            var_idx = int(model.cov[row, bi])
            knot_std = float(model.root[row, bi])
            # Back-transform knot to original scale
            knot = knot_std * X_std[var_idx] + X_mean[var_idx] if 0 <= var_idx < len(X_mean) else knot_std
            is_hinge = bool(model.truncated[row, bi])
            coef = float(model.coefficients[j]) if j < len(model.coefficients) else np.nan
            var_name = predictors[var_idx] if 0 <= var_idx < len(predictors) else f"var_{var_idx}"
            basis_info.append({
                'outcome': outcome,
                'variable': var_name,
                'knot': round(knot, 2),
                'type': 'hinge' if is_hinge else 'linear',
                'coefficient': round(coef, 4),
                'basis_index': int(bi)
            })

    # Store partial dependence data for plotting (back-transform x to original scale)
    partial_dep = {}
    for vi, vname in enumerate(predictors):
        if vname not in diastolic_vars:
            continue
        x_range_std = np.linspace(np.nanpercentile(X[:, vi], 2),
                                  np.nanpercentile(X[:, vi], 98), 200)
        X_grid = np.tile(np.median(X, axis=0), (200, 1))
        X_grid[:, vi] = x_range_std
        y_grid = model(X_grid)
        x_range_orig = x_range_std * X_std[vi] + X_mean[vi]
        partial_dep[vname] = pd.DataFrame({
            'x': x_range_orig, 'y_pred': y_grid,
            'variable': vname, 'outcome': outcome
        })

    omar_results[outcome] = {
        'basis': pd.DataFrame(basis_info) if basis_info else pd.DataFrame(),
        'r2': round(r2, 4),
        'lof': round(lof, 4),
        'n': len(df_complete),
        'predictors': predictors,
        'partial_dep': pd.concat(partial_dep.values()) if partial_dep else pd.DataFrame(),
        'X': df_complete[predictors],
        'y': df_complete[outcome]
    }

# Print discovered thresholds
for outcome, res in omar_results.items():
    label = "Peak VO₂ (FRIEND 2.0 %pred)" if "VO2" in outcome else "VE/VCO₂ Slope"
    print(f"\n{'='*60}")
    print(f"OMAR Threshold Discovery: {label}")
    print(f"  N = {res['n']}, R² = {res['r2']:.3f}, LOF = {res['lof']:.3f}")
    print(f"{'='*60}")
    if len(res['basis']) > 0:
        diastolic_hits = res['basis'][res['basis']['variable'].isin(diastolic_vars)]
        if len(diastolic_hits) > 0:
            print("\n  Discovered diastolic thresholds:")
            for _, row in diastolic_hits.iterrows():
                print(f"    {row['variable']}: knot = {row['knot']}, "
                      f"type = {row['type']}, coef = {row['coefficient']}")
        else:
            print("  No diastolic thresholds selected by OMAR.")

        other_hits = res['basis'][~res['basis']['variable'].isin(diastolic_vars)]
        if len(other_hits) > 0:
            print("\n  Other selected variables:")
            for _, row in other_hits.iterrows():
                print(f"    {row['variable']}: knot = {row['knot']}, "
                      f"type = {row['type']}, coef = {row['coefficient']}")
    else:
        print("  No basis functions extracted.")
```

```{r omar_export, results='asis'}
# Export OMAR thresholds from Python results
if (exists("omar_results", envir = reticulate::py)) {
  py_res <- reticulate::py$omar_results
  omar_thresholds <- bind_rows(lapply(names(py_res), function(outcome) {
    res <- py_res[[outcome]]
    bf_df <- res$basis
    if (!is.null(bf_df) && nrow(bf_df) > 0) {
      bf_df$r2  <- res$r2
      bf_df$lof <- res$lof
      bf_df$n   <- res$n
      bf_df
    }
  }))

  if (nrow(omar_thresholds) > 0) {
    # Diastolic thresholds table
    omar_diastolic <- omar_thresholds %>%
      filter(variable %in% c("e_e_ave", "la_vol_index", "tr_max_vel",
                               "e_prime_ave", "mv_e_a", "mv_dec_time")) %>%
      mutate(across(where(is.numeric), ~ round(., 3)))

    if (nrow(omar_diastolic) > 0) {
      omar_diastolic %>%
        flextable() %>%
        bold(part = "header") %>%
        theme_booktabs() %>%
        autofit() %>%
        set_caption("Table S. OMAR-Discovered Diastolic Thresholds for CPET Outcomes")
    }

    # Export to Supplemental folder
    dir.create("../Manuscript/Supplemental", showWarnings = FALSE, recursive = TRUE)
    write.csv(omar_thresholds,
              "../Manuscript/Supplemental/Table_S1_OMAR_Thresholds.csv", row.names = FALSE)

    # Compare OMAR knots to ASE 2025 cutoffs
    ase_cutoffs <- tibble(
      variable = c("e_e_ave", "la_vol_index", "tr_max_vel", "e_prime_ave", "mv_e_a", "mv_e_a"),
      ase_cutoff = c(14, 34, 280, 6.5, 0.8, 2.0),
      ase_direction = c(">", ">", ">", "<=", "<=", ">=")
    )

    omar_vs_ase <- omar_diastolic %>%
      left_join(ase_cutoffs, by = "variable", relationship = "many-to-many") %>%
      filter(!is.na(ase_cutoff))

    if (nrow(omar_vs_ase) > 0) {
      cat("\n\n**OMAR vs ASE 2025 Cutoff Comparison:**\n\n")
      omar_vs_ase %>%
        select(outcome, variable, omar_knot = knot, type, ase_cutoff, ase_direction) %>%
        flextable() %>%
        bold(part = "header") %>%
        theme_booktabs() %>%
        autofit() %>%
        set_caption("OMAR-Discovered vs ASE 2025 Guideline Thresholds")

      # Export to Supplemental
      write.csv(omar_vs_ase %>%
                  select(outcome, variable, omar_knot = knot, type, ase_cutoff, ase_direction),
                "../Manuscript/Supplemental/Table_S2_OMAR_vs_ASE.csv", row.names = FALSE)
    }
  }
}
```

```{r omar_figure, fig.width=6.9, fig.height=6.5}
# OMAR partial dependence plots with discovered thresholds vs ASE 2025 cutoffs
if (exists("omar_results", envir = reticulate::py)) {
  py_res <- reticulate::py$omar_results

  # Collect partial dependence curves and knot locations
  pd_all <- bind_rows(lapply(names(py_res), function(outcome) {
    res <- py_res[[outcome]]
    if (!is.null(res$partial_dep) && nrow(res$partial_dep) > 0) {
      res$partial_dep
    }
  }))

  knots_all <- bind_rows(lapply(names(py_res), function(outcome) {
    res <- py_res[[outcome]]
    if (!is.null(res$basis) && nrow(res$basis) > 0) {
      res$basis %>%
        filter(type == "hinge",
               variable %in% c("e_e_ave", "la_vol_index", "tr_max_vel",
                                "e_prime_ave", "mv_e_a", "mv_dec_time"))
    }
  }))

  # ASE 2025 reference cutoffs for overlay
  ase_ref <- tibble(
    variable = c("e_e_ave", "la_vol_index", "tr_max_vel", "e_prime_ave", "mv_e_a", "mv_e_a"),
    ase_cutoff = c(14, 34, 280, 6.5, 0.8, 2.0),
    label = c("ASE: >14", "ASE: >34", "ASE: >280", "ASE: \u22646.5", "ASE: \u22640.8", "ASE: \u22652.0")
  )

  # Variable display labels
  var_labels <- c(
    "e_e_ave"      = "E/e' (average)",
    "la_vol_index"  = "LAVI (mL/m\u00B2)",
    "tr_max_vel"    = "TR Vmax (cm/s)",
    "e_prime_ave"   = "Average e' (cm/s)",
    "mv_e_a"        = "E/A Ratio",
    "mv_dec_time"   = "MV Decel Time (ms)"
  )

  outcome_labels <- c(
    "VO2_FRIEND2_PP" = "Peak VO\u2082 (% Predicted)",
    "VeVco2_slope"   = "VE/VCO\u2082 Slope"
  )

  if (nrow(pd_all) > 0) {
    pd_all <- pd_all %>%
      mutate(var_label = var_labels[variable],
             outcome_label = outcome_labels[outcome])

    # Build figure: partial dependence curves with OMAR knots and ASE cutoffs
    fig_omar <- ggplot(pd_all, aes(x = x, y = y_pred)) +
      geom_line(color = jacc_cols["navy"], linewidth = 1) +
      # OMAR-discovered knots (solid red)
      {if (nrow(knots_all) > 0) {
        knots_plot <- knots_all %>%
          mutate(var_label = var_labels[variable],
                 outcome_label = outcome_labels[outcome])
        geom_vline(data = knots_plot,
                   aes(xintercept = knot),
                   color = jacc_cols["red"], linetype = "solid", linewidth = 0.6)
      }} +
      # ASE 2025 cutoffs (dashed blue)
      {
        ase_plot <- ase_ref %>%
          inner_join(pd_all %>% distinct(variable, outcome, var_label, outcome_label),
                     by = "variable", relationship = "many-to-many")
        if (nrow(ase_plot) > 0) {
          geom_vline(data = ase_plot,
                     aes(xintercept = ase_cutoff),
                     color = jacc_cols["blue"], linetype = "dashed", linewidth = 0.5)
        }
      } +
      facet_wrap(outcome_label ~ var_label, scales = "free", ncol = 3) +
      labs(title = "OMAR Partial Dependence: Diastolic Indices and CPET Outcomes",
           subtitle = "Solid red = OMAR-discovered knot | Dashed blue = ASE 2025 cutoff",
           x = NULL,
           y = "Predicted Outcome") +
      theme_jacc() +
      theme(strip.text = element_text(size = 7))

    print(fig_omar)
    save_jacc(fig_omar, "../2_Output/Figure_OMAR_Thresholds.pdf", w = 6.9, h = 6.5)
    save_jacc(fig_omar, "../Manuscript/Supplemental/Figure_S1_OMAR_Thresholds.pdf", w = 6.9, h = 6.5)
  }
}
```

## OMAR Methods

To complement guideline-based categorical grading with data-driven threshold discovery, we employed OMAR (Open Multivariate Adaptive Regression), a modern open-source implementation of the multivariate adaptive regression splines (MARS) algorithm (Walter et al., Technical University of Munich). OMAR uses a forward/backward stepwise procedure to automatically identify piecewise-linear breakpoints ("hinge functions") in the relationship between predictor variables and outcomes, without requiring pre-specified knot locations. In the forward pass, candidate basis functions are added at data-driven breakpoints where the slope of the predictor-outcome relationship changes; in the backward pass, basis functions that do not improve the generalized cross-validation (GCV) criterion are pruned, retaining only structurally important thresholds. Models were fit with a maximum of 15 basis functions and a GCV penalty of 3.0 to guard against overfitting. For each CPET outcome (Peak VO$_2$ [FRIEND 2.0 % predicted] and VE/VCO$_2$ slope), diastolic indices (E/e$'$, LAVI, TR V$_{max}$, average e$'$, E/A ratio, deceleration time) were included as candidate predictors alongside age, lean body mass index, and sex. Partial dependence curves were computed for each selected diastolic variable by varying it across its observed range while holding all other predictors at their median values. OMAR-discovered knots were compared against ASE 2025 guideline cutoffs to assess concordance between data-driven and expert-derived thresholds.

## OMAR Results

OMAR identified data-driven breakpoints in the relationship between diastolic indices and cardiopulmonary exercise capacity. Among the diastolic variables entered as candidates, OMAR retained those with structurally significant piecewise-linear associations, automatically discarding variables that did not improve model fit beyond the GCV penalty. Partial dependence plots illustrate the shape of these relationships, with OMAR-discovered knots (solid red lines) compared against corresponding ASE 2025 guideline cutoffs (dashed blue lines). Concordance between data-driven and guideline-derived thresholds supports the clinical validity of the ASE 2025 cutoffs in this HCM population; discordance identifies parameters where HCM-specific thresholds may warrant further investigation.

*[Specific knot values and concordance/discordance findings to be reported after rendering.]*

# Longitudinal GAMM Analysis

We employ a Generalized Additive Mixed Model (GAMM) via `mgcv::bam()` to model the nonlinear trajectory of Peak VO$_2$ over time, testing whether the trajectory diverges as a function of baseline diastolic severity via a tensor product interaction.

```{r gamm_analysis, fig.width=6.9, fig.height=5.5}
# Derive baseline DD composite Z-score from GAMLSS (if available) or standardized E/e'
if ("e_e_ave" %in% names(gamlss_results)) {
  # Use SEAMLSS Z-score from E/e' model
  z_source <- gamlss_results[["e_e_ave"]]$data %>%
    select(ID, dd_zscore = seamlss_z)
  long_df <- long_df %>%
    left_join(z_source, by = "ID")
} else {
  # Fallback: standardized E/e'
  long_df <- long_df %>%
    mutate(dd_zscore = scale(as.numeric(as.character(e_e_ave)))[, 1])
}

# Filter to patients with DD Z-score and sufficient follow-up
gamm_df <- long_df %>%
  filter(!is.na(VO2_FRIEND2_PP), !is.na(dd_zscore), !is.na(time_yrs),
         !is.na(age), !is.na(LBMI), is.finite(time_yrs)) %>%
  mutate(ID_fac = factor(ID))

cat(sprintf("GAMM cohort: %d observations, %d patients\n",
            nrow(gamm_df), n_distinct(gamm_df$ID)))

if (nrow(gamm_df) >= 50 && n_distinct(gamm_df$ID) >= 20) {
  # Fit GAMM with tensor product interaction
  gamm_fit <- tryCatch({
    bam(VO2_FRIEND2_PP ~ s(time_yrs, bs = "cr", k = 5) +
          s(dd_zscore, bs = "cr", k = 5) +
          ti(time_yrs, dd_zscore, k = c(4, 4)) +
          age + Sex + LBMI +
          s(ID_fac, bs = "re"),
        data = gamm_df, method = "fREML", discrete = TRUE)
  }, error = function(e) {
    message("GAMM failed, fitting simpler model: ", e$message)
    bam(VO2_FRIEND2_PP ~ s(time_yrs, bs = "cr", k = 5) +
          dd_zscore * time_yrs +
          age + Sex + LBMI +
          s(ID_fac, bs = "re"),
        data = gamm_df, method = "fREML", discrete = TRUE)
  })

  # Comparison: linear mixed model
  lme_fit <- lmer(VO2_FRIEND2_PP ~ time_yrs * dd_zscore + age + Sex + LBMI + (1 | ID),
                  data = gamm_df, REML = FALSE)

  cat(sprintf("GAMM AIC: %.1f  |  LME AIC: %.1f\n", AIC(gamm_fit), AIC(lme_fit)))

  # Summary of smooth terms
  gamm_summary <- summary(gamm_fit)
  cat("\n**GAMM Smooth Terms:**\n")
  print(gamm_summary$s.table)

  write.csv(as.data.frame(gamm_summary$s.table),
            "../2_Output/Stats_GAMM_SmoothTerms.csv")
  write.csv(as.data.frame(gamm_summary$p.table),
            "../2_Output/Stats_GAMM_ParametricTerms.csv")

  # Prediction: trajectories stratified by DD Z-score tertiles
  z_tertiles <- quantile(gamm_df$dd_zscore, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)

  pred_gamm <- expand_grid(
    time_yrs  = seq(0, min(5, max(gamm_df$time_yrs, na.rm = TRUE)), by = 0.25),
    dd_zscore = z_tertiles
  ) %>%
    mutate(
      level = factor(dd_zscore, labels = c("Q1 (less impaired)", "Median", "Q3 (more impaired)")),
      age  = median(gamm_df$age, na.rm = TRUE),
      Sex  = factor(names(which.max(table(gamm_df$Sex))), levels = levels(gamm_df$Sex)),
      LBMI = median(gamm_df$LBMI, na.rm = TRUE),
      ID_fac = gamm_df$ID_fac[1]
    )

  pred_gamm$fit <- predict(gamm_fit, newdata = pred_gamm, exclude = "s(ID_fac)", type = "response")

  fig4 <- ggplot(pred_gamm, aes(x = time_yrs, y = fit, color = level)) +
    geom_line(linewidth = 1.1) +
    scale_color_manual(values = c(jacc_cols["blue"], jacc_cols["navy"], jacc_cols["red"])) +
    labs(title = bquote(bold("Longitudinal Peak" ~ VO[2] ~ "Trajectories by Baseline DD Severity")),
         subtitle = "GAMM with tensor product interaction: time \u00D7 DD Z-score",
         x = "Years from Baseline CPET",
         y = bquote("Predicted Peak" ~ VO[2] ~ "(FRIEND 2.0 % Pred)"),
         color = "Baseline DD Z-score") +
    theme_jacc() +
    theme(legend.position = "right")

  fig4
  save_jacc(fig4, "../2_Output/Figure4_GAMM_Trajectories.pdf", w = 6.9, h = 5.0)
}
```

# Heart Failure Outcome Analysis

Time-to-event analysis tests whether ASE 2025 filling pressure classification (Normal vs Elevated) and continuous DD Z-scores predict a composite heart failure endpoint (acute HF, chronic HF, transplant, or death).

```{r hf_outcomes, fig.width=6.9, fig.height=5.0, results='asis'}
surv_df <- baseline_df %>%
  filter(!is.na(hf_composite), !is.na(hf_composite_yrs) | hf_composite == 0) %>%
  mutate(
    # For censored patients, use time from CPX to last encounter
    follow_up_yrs = ifelse(
      hf_composite == 1 & !is.na(hf_composite_yrs),
      hf_composite_yrs,
      as.numeric(as.Date(last_enc_date) - cpx_test_date) / 365.25
    ),
    follow_up_yrs = ifelse(is.na(follow_up_yrs) | follow_up_yrs <= 0, 0.01, follow_up_yrs)
  ) %>%
  filter(!is.na(follow_up_yrs), follow_up_yrs > 0)

cat(sprintf("Survival cohort: %d patients, %d events\n",
            nrow(surv_df), sum(surv_df$hf_composite == 1, na.rm = TRUE)))

if (nrow(surv_df) >= 30 && sum(surv_df$hf_composite == 1, na.rm = TRUE) >= 5) {

  # Kaplan-Meier by filling pressure classification
  surv_fp <- surv_df %>% filter(fp_class != "Indeterminate", !is.na(fp_class))

  if (nrow(surv_fp) >= 20) {
    km_fit <- survfit(Surv(follow_up_yrs, hf_composite) ~ fp_class, data = surv_fp)

    # KM plot
    km_data <- broom::tidy(km_fit) %>%
      mutate(strata = str_remove(strata, "fp_class="))

    fig5 <- ggplot(km_data, aes(x = time, y = estimate, color = strata)) +
      geom_step(linewidth = 0.9) +
      geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = strata),
                  alpha = 0.15, stat = "identity") +
      scale_color_manual(values = c("Normal" = jacc_cols["blue"],
                                     "Elevated" = jacc_cols["red"])) +
      scale_fill_manual(values = c("Normal" = jacc_cols["blue"],
                                    "Elevated" = jacc_cols["red"])) +
      scale_y_continuous(limits = c(0, 1), labels = percent) +
      labs(title = "Composite HF Endpoint by LV Filling Pressure",
           subtitle = "ASE 2025 HCM-specific algorithm",
           x = "Years from Baseline CPET",
           y = "Event-Free Survival",
           color = "Filling Pressure", fill = "Filling Pressure") +
      theme_jacc() +
      theme(legend.position = "right")

    fig5
    save_jacc(fig5, "../2_Output/Figure5_KM_FillingPressure.pdf", w = 6.9, h = 5.0)

    # Log-rank test
    lr_test <- survdiff(Surv(follow_up_yrs, hf_composite) ~ fp_class, data = surv_fp)
    cat(sprintf("\n**Log-rank test p-value: %.4f**\n\n", 1 - pchisq(lr_test$chisq, df = length(lr_test$n) - 1)))
  }

  # Cox proportional hazards: Filling pressure (binary)
  cox_fp <- tryCatch({
    coxph(Surv(follow_up_yrs, hf_composite) ~ fp_class + age + Sex + LBMI,
          data = surv_fp)
  }, error = function(e) NULL)

  if (!is.null(cox_fp)) {
    cat("\n**Cox Model: Filling Pressure Classification**\n")
    tidy(cox_fp, exponentiate = TRUE, conf.int = TRUE) %>%
      mutate(across(where(is.numeric), ~ round(., 3))) %>%
      flextable() %>%
      bold(part = "header") %>%
      theme_booktabs() %>%
      autofit() %>%
      set_caption("Table 3. Cox Proportional Hazards: Composite HF Endpoint") %>%
      print()

    write.csv(tidy(cox_fp, exponentiate = TRUE, conf.int = TRUE),
              "../2_Output/Table3_Cox_FillingPressure.csv", row.names = FALSE)

    # PH assumption check
    ph_check <- cox.zph(cox_fp)
    cat(sprintf("\n**Proportional hazards global test p = %.3f**\n", ph_check$table["GLOBAL", "p"]))
  }

  # Cox: continuous DD Z-score (from GAMLSS if available)
  if ("e_e_ave" %in% names(gamlss_results)) {
    z_src <- gamlss_results[["e_e_ave"]]$data %>% select(ID, dd_zscore = seamlss_z)
    surv_z <- surv_df %>% left_join(z_src, by = "ID") %>% filter(!is.na(dd_zscore))

    if (nrow(surv_z) >= 20 && sum(surv_z$hf_composite == 1, na.rm = TRUE) >= 5) {
      cox_z <- coxph(Surv(follow_up_yrs, hf_composite) ~ dd_zscore + age + Sex + LBMI,
                     data = surv_z)
      cat("\n**Cox Model: Continuous DD Z-Score**\n")
      tidy(cox_z, exponentiate = TRUE, conf.int = TRUE) %>%
        mutate(across(where(is.numeric), ~ round(., 3))) %>%
        flextable() %>% bold(part = "header") %>% theme_booktabs() %>% autofit() %>% print()

      write.csv(tidy(cox_z, exponentiate = TRUE, conf.int = TRUE),
                "../2_Output/Table3_Cox_DDZscore.csv", row.names = FALSE)
    }
  }
} else {
  cat("Insufficient events for survival analysis.\n")
}
```

# Patient Characteristics

## Table 1

```{r table1, results='asis'}
table1_data <- baseline_df %>%
  mutate(
    VeVco2_slope = as.numeric(as.character(VeVco2_slope)),
    HCM_Phenotype = as.factor(HCM_Phenotype)
  ) %>%
  select(fp_class, age, Sex, BMI, BSA, LBMI, HCM_Phenotype,
         pk.RER, VO2_FRIEND2_PP, VO2_WASSERMAN_PP, VeVco2_slope, HRR,
         e_prime_ave, med_peak_e_vel, lat_peak_e_vel,
         e_e_ave, la_vol_index, E_vel, tr_max_vel, mv_e_a, mv_dec_time, ef_modsp4)

# Overall
table1_overall <- table1_data %>%
  select(-fp_class) %>%
  tbl_summary(
    label = list(
      age              ~ "Age, years",
      Sex              ~ "Sex",
      BMI              ~ "Body Mass Index, kg/m\u00B2",
      BSA              ~ "Body Surface Area, m\u00B2",
      LBMI             ~ "Lean Body Mass Index, kg/m\u00B2",
      HCM_Phenotype    ~ "HCM Phenotype",
      pk.RER           ~ "Peak RER",
      VO2_FRIEND2_PP   ~ "Peak VO\u2082 (FRIEND 2.0 %pred)",
      VO2_WASSERMAN_PP ~ "Peak VO\u2082 (Wasserman %pred)",
      VeVco2_slope     ~ "VE/VCO\u2082 Slope",
      HRR              ~ "Heart Rate Recovery (1 min)",
      e_prime_ave      ~ "Average e', cm/s",
      med_peak_e_vel   ~ "Septal e', cm/s",
      lat_peak_e_vel   ~ "Lateral e', cm/s",
      e_e_ave          ~ "E/e' (average)",
      la_vol_index     ~ "LAVI, mL/m\u00B2",
      E_vel            ~ "Mitral E Velocity, cm/s",
      tr_max_vel       ~ "TR Vmax, cm/s",
      mv_e_a           ~ "E/A Ratio",
      mv_dec_time      ~ "MV Deceleration Time, ms",
      ef_modsp4        ~ "LVEF, %"
    ),
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 1,
    missing = "no"
  ) %>%
  bold_labels() %>%
  modify_caption("**Table 1. Baseline Patient Characteristics**")

table1_overall

table1_overall %>%
  as_flex_table() %>%
  save_as_docx(path = "../2_Output/Table1_Manuscript.docx")
```

## Figure 1: Cohort Characteristics

```{r figure1, fig.width=6.9, fig.height=7.5}
accent  <- jacc_cols["blue"]
gray_f  <- "grey78"

# Panel A: Age
p_age <- ggplot(table1_data, aes(x = age)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 5, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = accent, linewidth = 0.8) +
  labs(title = "Age", x = "Years", y = "Density") +
  theme_jacc()

# Panel B: BMI
p_bmi <- ggplot(table1_data, aes(x = BMI)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 2, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = accent, linewidth = 0.8) +
  labs(title = "BMI", x = bquote("kg/" * m^2), y = "Density") +
  theme_jacc()

# Panel C: LBMI
p_lbmi <- ggplot(table1_data, aes(x = LBMI)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = accent, linewidth = 0.8) +
  labs(title = "LBMI", x = bquote("kg/" * m^2), y = "Density") +
  theme_jacc()

# Panel D: Peak VO2 comparison
vo2_long <- table1_data %>%
  select(VO2_FRIEND2_PP, VO2_WASSERMAN_PP) %>%
  pivot_longer(everything(), names_to = "Equation", values_to = "Value") %>%
  mutate(Equation = recode(Equation,
                            "VO2_FRIEND2_PP" = "FRIEND 2.0",
                            "VO2_WASSERMAN_PP" = "Wasserman"))

p_vo2 <- ggplot(vo2_long, aes(x = Value, fill = Equation)) +
  geom_histogram(aes(y = after_stat(density)), bins = 25, alpha = 0.35, position = "identity", color = "white") +
  geom_density(aes(color = Equation), linewidth = 0.6) +
  scale_fill_manual(values = c("FRIEND 2.0" = jacc_cols["blue"], "Wasserman" = jacc_cols["gray"])) +
  scale_color_manual(values = c("FRIEND 2.0" = jacc_cols["navy"], "Wasserman" = jacc_cols["gray"])) +
  scale_x_continuous(limits = c(0, 200)) +
  labs(title = bquote(bold("Peak" ~ VO[2] ~ "Comparison")),
       x = bquote("% Predicted" ~ VO[2]), y = "Density") +
  theme_jacc()

# Panel E: VE/VCO2
p_vevco2 <- ggplot(table1_data, aes(x = VeVco2_slope)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = accent, linewidth = 0.8) +
  labs(title = bquote(bold(VE/VCO[2] ~ "Slope")), x = bquote(VE/VCO[2]), y = "Density") +
  theme_jacc()

# Panel F-I: Echo distributions
p_ee <- ggplot(table1_data, aes(x = e_e_ave)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = jacc_cols["red"], linewidth = 0.8) +
  geom_vline(xintercept = 14, linetype = "dashed", color = jacc_cols["red"], linewidth = 0.4) +
  labs(title = "E/e' (average)", x = "E/e'", y = "Density") +
  theme_jacc()

p_lavi <- ggplot(table1_data, aes(x = la_vol_index)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = jacc_cols["red"], linewidth = 0.8) +
  geom_vline(xintercept = 34, linetype = "dashed", color = jacc_cols["red"], linewidth = 0.4) +
  labs(title = "LAVI", x = bquote("mL/" * m^2), y = "Density") +
  theme_jacc()

p_trv <- ggplot(table1_data, aes(x = tr_max_vel)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = jacc_cols["red"], linewidth = 0.8) +
  geom_vline(xintercept = 280, linetype = "dashed", color = jacc_cols["red"], linewidth = 0.4) +
  labs(title = "TR Vmax", x = "cm/s", y = "Density") +
  theme_jacc()

p_eprime <- ggplot(table1_data, aes(x = e_prime_ave)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = jacc_cols["red"], linewidth = 0.8) +
  geom_vline(xintercept = 6.5, linetype = "dashed", color = jacc_cols["red"], linewidth = 0.4) +
  labs(title = "Average e'", x = "cm/s", y = "Density") +
  theme_jacc()

p_ea <- ggplot(table1_data, aes(x = mv_e_a)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = jacc_cols["red"], linewidth = 0.8) +
  geom_vline(xintercept = c(0.8, 2.0), linetype = "dashed", color = jacc_cols["red"], linewidth = 0.4) +
  labs(title = "E/A Ratio", x = "E/A", y = "Density") +
  theme_jacc()

p_ef <- ggplot(table1_data, aes(x = ef_modsp4)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = gray_f, alpha = 0.7, color = "white") +
  geom_density(color = accent, linewidth = 0.8) +
  labs(title = "LVEF", x = "%", y = "Density") +
  theme_jacc()

fig1 <- (p_age | p_bmi | p_lbmi) /
        (p_vo2 | p_vevco2 | p_eprime) /
        (p_ee | p_lavi | p_trv) /
        (p_ea | p_ef | plot_spacer()) +
  plot_annotation(
    tag_levels = "A",
    title = "Baseline Cohort Biometrics, Functional Capacity, and Echocardiography"
  ) &
  theme(plot.tag = element_text(face = "bold", size = 10))

fig1
save_jacc(fig1, "../2_Output/Figure1_Cohort_Characteristics.pdf", w = 6.9, h = 9.5)
```

## Figure 2: Longitudinal Analysis

```{r figure2, fig.width=6.9, fig.height=5.5}
# Annualized slopes
CPX_slopes <- cpx_echo %>%
  mutate(VO2_FRIEND2_PP = as.numeric(as.character(VO2_FRIEND2_PP)),
         VeVco2_slope = as.numeric(as.character(VeVco2_slope))) %>%
  group_by(ID) %>%
  arrange(days_from_baseline) %>%
  summarize(
    follow_up_years    = max(days_from_baseline) / 365.25,
    annual_delta_vo2   = (last(VO2_FRIEND2_PP) - first(VO2_FRIEND2_PP)) / (max(days_from_baseline) / 365.25),
    baseline_vo2       = first(VO2_FRIEND2_PP),
    baseline_vevco2    = first(VeVco2_slope),
    num_tests          = n(),
    .groups = "drop"
  ) %>%
  filter(num_tests > 1, is.finite(annual_delta_vo2))

# Panel A: Spaghetti plot
p_spaghetti <- ggplot(cpx_echo %>% mutate(VO2_FRIEND2_PP = as.numeric(as.character(VO2_FRIEND2_PP))),
                       aes(x = days_from_baseline / 365.25, y = VO2_FRIEND2_PP)) +
  geom_line(aes(group = ID), alpha = 0.15, color = jacc_cols["navy"]) +
  geom_point(alpha = 0.25, size = 0.8, color = jacc_cols["blue"]) +
  geom_smooth(method = "lm", color = jacc_cols["red"], linewidth = 1, se = TRUE,
              fill = jacc_cols["red"], alpha = 0.1) +
  labs(title = bquote(bold("Peak" ~ VO[2] ~ "Trajectories")),
       x = "Years from Baseline", y = bquote("Peak" ~ VO[2] ~ "(% Pred)")) +
  theme_jacc()

# Panel B: Swimmer plot
p_swimmer <- ggplot(cpx_echo, aes(x = days_from_baseline / 365.25, y = as.factor(ID))) +
  geom_line(aes(group = ID), color = jacc_cols["navy"], linewidth = 0.3) +
  geom_point(aes(color = as.numeric(as.character(VO2_FRIEND2_PP))), size = 1) +
  scale_color_viridis_c(option = "magma", name = "% Pred") +
  labs(title = "Patient Timeline", x = "Years Since Baseline", y = NULL) +
  theme_jacc() + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

# Panel C: Correlation heatmap
corr_data <- CPX_slopes %>%
  left_join(table1_data %>% select(any_of(c("ID", "age", "BMI", "LBMI"))),
            by = character()) %>%  # will use table1_data from baseline
  select(any_of(c("baseline_vo2", "baseline_vevco2", "annual_delta_vo2", "follow_up_years")))

# Use baseline_df for correlation
corr_src <- baseline_df %>%
  left_join(CPX_slopes %>% select(ID, annual_delta_vo2, follow_up_years), by = "ID") %>%
  select(age, BMI, LBMI,
         `Baseline VO2%` = VO2_FRIEND2_PP,
         `VE/VCO2` = VeVco2_slope,
         `Annual dVO2%` = annual_delta_vo2,
         `Follow-up (Yrs)` = follow_up_years) %>%
  mutate(across(everything(), ~ as.numeric(as.character(.)))) %>%
  drop_na()

if (nrow(corr_src) >= 10) {
  corr_matrix <- round(cor(corr_src, use = "pairwise.complete.obs"), 2)
  p_corr <- ggcorrplot(corr_matrix, hc.order = TRUE, type = "lower", lab = TRUE,
                        lab_size = 2.5,
                        colors = c(jacc_cols["red"], "white", jacc_cols["green"]),
                        title = "Parameter Correlations") +
    theme_jacc() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
          axis.text.y = element_text(size = 7))
} else {
  p_corr <- ggplot() + annotate("text", x = 0.5, y = 0.5, label = "Insufficient data") + theme_void()
}

fig2 <- (p_spaghetti | p_swimmer | p_corr) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "bold", size = 11))

fig2
save_jacc(fig2, "../2_Output/Figure2_Longitudinal_Analysis.pdf", w = 6.9, h = 5.5)
```

# Supplemental Analyses

## Age-residualized diastolic indices

```{r residualize, fig.width=6.9, fig.height=4.5}
baseline_df_resid <- baseline_df
for (idx in diastolic_indices) {
  baseline_df_resid <- residualize_on_age(baseline_df_resid, idx)
}

resid_summary <- bind_rows(lapply(diastolic_indices, function(idx) {
  resid_col <- paste0(idx, "_resid")
  tibble(
    index    = idx,
    label    = label_map[idx],
    n_raw    = sum(!is.na(baseline_df[[idx]])),
    n_resid  = sum(!is.na(baseline_df_resid[[resid_col]])),
    sd_resid = round(sd(baseline_df_resid[[resid_col]], na.rm = TRUE), 3)
  )
}))

resid_summary %>%
  flextable() %>%
  bold(part = "header") %>%
  theme_booktabs() %>%
  autofit() %>%
  set_caption("Table S. Age-Residualized Diastolic Indices Summary")

# Diagnostic boxplot
resid_long <- baseline_df_resid %>%
  select(all_of(paste0(diastolic_indices, "_resid"))) %>%
  pivot_longer(everything(), names_to = "index", values_to = "resid") %>%
  mutate(index = str_replace(index, "_resid$", ""),
         label = label_map[index])

p_resid <- ggplot(resid_long, aes(x = label, y = resid)) +
  geom_boxplot(fill = "grey85", color = "black", outlier.size = 0.5, linewidth = 0.3) +
  coord_flip() +
  labs(title = "Age-Residualized Diastolic Indices", x = NULL, y = "Residual") +
  theme_jacc()

p_resid
save_jacc(p_resid, "../2_Output/FigureS_AgeResidualized_DiastolicIndices.pdf", w = 6.9, h = 4.5)
```

## Export missing echo audit

```{r export_missing_echo}
echo_vars_present <- intersect(
  c("e_e_ave", "e_e_lat", "e_e_med", "med_peak_e_vel", "ef_modsp4",
    "la_vol", "la_vol_index", "lat_peak_e_vel", "mv_a_point", "mv_a_dur",
    "mv_dec_time", "mv_e_a", "pulm_dias_vel", "tr_max_vel"),
  colnames(cpx_echo)
)

missing_echo <- cpx_aligned %>%
  anti_join(cpx_echo %>% select(ID, cpx_test_date) %>% distinct(),
            by = c("ID", "cpx_test_date")) %>%
  arrange(MRN, cpx_test_date) %>%
  select(ID, MRN, CPX.sequential, cpx_test_date, days_from_baseline, age, Sex, BMI, LBMI,
         VO2_FRIEND2_PP, VeVco2_slope)

if (nrow(missing_echo) > 0) {
  openxlsx::write.xlsx(
    list(Missing_Echo_CPX = missing_echo,
         Summary = tibble(n_missing = nrow(missing_echo),
                          n_patients = n_distinct(missing_echo$ID))),
    file = "../2_Output/Serial_CPX_missing_resting_echo.xlsx",
    overwrite = TRUE
  )
  cat(sprintf("Exported %d CPX tests without echo linkage (%d patients)\n",
              nrow(missing_echo), n_distinct(missing_echo$ID)))
}
```

```{r session_info}
cat("\n\n---\n**Session Info:**\n")
sessionInfo()
```
